---
title: "Abud vs Villani, after combat - sample to sample correlation"
author: "Izaskun Mallona, Mark D. Robinson's lab, UZH"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: kable
    theme: lumen
params:
  seed: 665
---

This adapts 03_custom/01_custom_analysis_postmeeting.Rmd to get a sample-to-sample (Spearman_ correlation plot of Villani vs Abud, including mutants, after combat correction, and on all genes (not only pluripotecy ones), to showcase iMGLs vs IPSCs transcriptomic closeness


## Diagnosis

```{r warning = FALSE, message = FALSE}
## .libPaths('/home/imallona/R/R4_bioc314')

library(ggplot2)
library(SingleCellExperiment)
## library(corrplot)
## library(recount3)
library(ComplexHeatmap)
## library(circlize)
library(dplyr)
library(sva)
library(limma)
library(ggrepel)

```

```{r}
## render on error
knitr::knit_hooks$set(error = function(x, options) {
    knitr::knit_exit()  
})
```

```{r}
knitr::opts_chunk$set(fig.width = 5,
                      fig.height = 5,
                      cache = TRUE,
                      include = TRUE,
                      fig.path = "plots_post/",
                      dev = c("png", "svg"),
                      cache.lazy = FALSE,
                      warning = TRUE,
                      message = TRUE)

## plan("multiprocess", workers = NTHREADS)
## options(future.globals.maxSize= 1.4e9)

## options(bitmapType='cairo')
```



```{r}
tmp3 <- readRDS('../03_custom/villani_plus_abud_sce.rds')
```

Original MDS (batch corrected)

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
mds <- reducedDim(tmp3)
mds$names <- tmp3$names
mds <- merge(mds, colData(tmp3), by = 'names')
rownames(mds) <- mds$names

print(ggplot(mds, aes(x=MDS1, y=MDS2, color = treatment, shape = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1"),
          y = sprintf("Dim. 2"),
          title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(legend.position='right') +
      guides(fill=guide_legend(ncol=2)) +
      theme(aspect.ratio=1) +
     scale_color_brewer(palette = "Set3"))

```

Repeat from scratch


```{r}

logcpms <- assay(tmp3, "logcpms")
# parametric adjustment
combat_abud <- sva::ComBat(dat=logcpms,
                           batch = tmp3$origin,
                           mod = NULL, par.prior = TRUE, prior.plots = TRUE)

```


```{r, fig.width = 8, fig.height = 5}
mds <- limma::plotMDS(combat_abud, top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = 7,
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)


vexp1 <- round(mds$var.explained[1]*100)
vexp2 <- round(mds$var.explained[2]*100)


if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = rownames(mds$distance.matrix.squared),
                      MDS1 = mds$x,
                      MDS2 = mds$y)
}

## mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'Abud', no = 'Ambra')

## mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

mds <- merge(mds, colData(tmp3), by = 'names')
rownames(mds) <- mds$names
```


```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", vexp1),
          y = sprintf("Dim. 2, var. exp. %s%%", vexp2),
          title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(aspect.ratio=1) +
      theme(legend.position='right')) #, legend.box='vertical', legend.text=element_text(size=9), legend.title=element_text(size=11),
                                        #axis.text=element_text(size=10), axis.title=element_text(size=12), title=element_text(size=12)))
```


Let's name the items to understand the reqs

```{r, fig.width = 10, fig.height = 10}
rownames(mds) -> mds$id


ggplot(mds, aes(x = MDS1, y = MDS2, color = treatment)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = treatment), size = 3, max.overlaps = Inf) +
  labs(
    x = sprintf("Dim. 1, var. exp. %s%%", vexp1),
    y = sprintf("Dim. 2, var. exp. %s%%", vexp2),
    title = "MDS after ComBat"
  ) +
  coord_cartesian(
    xlim = range(mds$MDS1, na.rm = TRUE),
    ylim = range(mds$MDS2, na.rm = TRUE),
    expand = TRUE
  ) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    legend.position = "right"
  )

```

So the correlations we're are interested in are:
- upperleftcorner: the iPSC_(Abud)
- middleleft: the iPSC_MG_(Abud)
- middleright: iPSC_MG_(Villani)

```{r}
## library(SingleCellExperiment)
## mat <- assay(tmp3, "logcpms")
mat <- combat_abud

cor_mat <- cor(mat, method = "spearman")

dim(cor_mat)

meta <- as.data.frame(colData(tmp3))

## ann_col <- HeatmapAnnotation( df = meta %>% select(treatment) )
ann_col <- HeatmapAnnotation(
  df = meta %>% select(treatment, origin, age)
)

## ann_row <- rowAnnotation( df = meta %>% select(origin) )
```

```{r, fig.width = 12, fig.height = 10}
cor_text_fun <- function(j, i, x, y, width, height, fill) {
  grid.text(sprintf("%.2f", cor_mat[i, j]),
            x = x, y = y,
            gp = gpar(fontsize = 8, col = "black"))
}

## col_fun <- colorRamp2(
##   c(0, 1),
##   c("white", "darkblue")
## )

col_fun <- colorRampPalette(c("white", "darkblue"))(100)



Heatmap(
  cor_mat,
  name = "Spearman\ncorrelation",
  col =  col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  top_annotation = ann_col,
  ## left_annotation = ann_row,
  ## cell_fun = cor_text_fun,
  rect_gp = gpar(col = "grey90", lwd = 0.5),
  heatmap_legend_param = list(title = "Correlation")
)

```

But the actual targets are:

- upperleftcorner: the iPSC_(Abud)
- middleleft: the iPSC_MG_(Abud)
- middleright: iPSC_MG_(Villani)

As stated by:

```
This indeed makes a lot of sense and would be an excellent way to address this point. Are you suggesting comparing not only Villani and Abud iMGLs, but also Villani iMGLs with iPSCs?
```


```{r, fig.width = 12, fig.height = 10}
cd <- colData(tmp3)
selected <- rownames(cd[cd$treatment %in% c('iPSC_(Abud)', 'iPSC_MG', 'iPSC_MG_(Abud)'),])

Heatmap(
  cor_mat[selected,selected],
  name = "Spearman\ncorrelation",
  col =  col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_dend = TRUE,
  show_column_dend = TRUE,
  top_annotation = ann_col[selected,],
  ## left_annotation = ann_row,
  ## cell_fun = cor_text_fun,
  rect_gp = gpar(col = "grey90", lwd = 0.5),
  heatmap_legend_param = list(title = "Correlation")
)

```

```{r, fig.width = 6, fig.height = 3.4}
cd <- as.data.frame(colData(tmp3))
selected <- rownames(cd[cd$treatment %in% c("iPSC_(Abud)", "iPSC_MG", "iPSC_MG_(Abud)"), ])

## subset cors and metadata
sub_cor <- cor_mat[selected, selected]
sub_meta <- droplevels(cd[selected, ])

cor_long <- data.frame(
  sample1 = rep(rownames(sub_cor), times = ncol(sub_cor)),
  sample2 = rep(colnames(sub_cor), each = nrow(sub_cor)),
  cor     = as.vector(sub_cor),
  stringsAsFactors = FALSE
)

# unique pairs only (i < j)
cor_long <- cor_long[cor_long$sample1 < cor_long$sample2, ]

cor_long$treatment1 <- sub_meta[cor_long$sample1, "treatment"]
cor_long$treatment2 <- sub_meta[cor_long$sample2, "treatment"]

cor_long$pair_group <- ifelse(
  cor_long$treatment1 == cor_long$treatment2,
  paste0("within_", cor_long$treatment1),
  paste(cor_long$treatment1, cor_long$treatment2, sep = " vs ")
)

(medians <- aggregate(cor ~ pair_group, data = cor_long, median))

## ggplot(cor_long, aes(x = cor, color = pair_group, fill = pair_group)) +
##   geom_density(alpha = 0.25) +
##     geom_vline(data = medians, aes(xintercept = cor, color = pair_group),
##                linetype = "dashed", linewidth = 0.7) +
##   theme_bw() +
##   labs(
##     title = "Pairwise transcriptome similarity",
##     x = "Spearman's correlation coefficient",
##     y = "Density"
##   )



ggplot(cor_long, aes(x = cor, color = pair_group, fill = pair_group)) +
  geom_density(alpha = 0.25) +

  geom_vline(
    data = medians,
    aes(xintercept = cor, color = pair_group),
    linetype = "dashed",
    linewidth = 0.7
  ) +

  ## geom_text_repel(
  ##   data = medians,
  ##   aes(x = cor, y = Inf, label = sprintf("%.2f", cor), color = pair_group),
  ##   direction = "y",
  ##   nudge_y = 0.1,
  ##   segment.color = NA,
  ##   size = 3,
  ##   force = 5,
  ##   max.overlaps = Inf
  ## ) +
  geom_label_repel(
      data = medians,
      aes(x = cor, y = Inf,
          label = sprintf("%.2f", cor),
          fill = pair_group),
      color = "black",
      nudge_y = 0.1,
      direction = "y",
      segment.color = NA,
      size = 3,
      label.size = 0.2, label.r = unit(0.1, "lines")) +

  coord_cartesian(clip = "off") +
  theme_bw() +
  labs(
    title = "Pairwise transcriptome similarity",
    x = "Spearman's correlation coefficient",
    y = "Density"
  )


```

# Timestamp

```{r sessionInfo2, cache = FALSE}
date()
sessionInfo()
## devtools::session_info()
```
