---
title: "Ambra Villani's microglia RNA-seq"
author: "Izaskun Mallona, Mark D. Robinson's lab, UZH"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: kable
    theme: lumen
params:
  seed: 665
---


```{r warning = FALSE, message = FALSE}
## .libPaths('/home/imallona/R/R4_bioc314')

library(ggplot2)
## library(EnhancedVolcano)
## library(iSEE)
## library(knitr)
## library(DRIMSeq)
## library(EnsDb.Hsapiens.v86)
library(SingleCellExperiment)
library(corrplot)
#library(GGally)
#library(scuttle)
#library(knitr)
#library(greybox) # cramer's distance
#library(viridis)
#library(regioneR)
#library(GenomicRanges)
#library(rtracklayer)

## camera start

## library(dplyr)
## library(msigdbr)
## library(CAMERA)

#require(topGO)
#require(org.Hs.eg.db)

## library("ggVennDiagram")
library(recount3)

 library(tibble)
library(dplyr)
##  library(tidyr)
library(limma)
library(edgeR)
library(sva)
library(DT)
## library(reshape2)


```


```{r edgeR-load-pkg}
suppressPackageStartupMessages({
    library(dplyr)
    library(tximport)
    library(tximeta)
    library(SingleCellExperiment)
    library(edgeR)
    library(ggplot2)
    library(msigdbr)
    library(EnhancedVolcano)
})
```



```{r}
## render on error
knitr::knit_hooks$set(error = function(x, options) {
    knitr::knit_exit()  
})
```

```{r}
knitr::opts_chunk$set(fig.width = 8,
                      fig.height = 8,
                      cache = TRUE,
                      include = TRUE,
                      fig.path = "plots_post/",
                      dev = c("png", "svg"),
                      cache.lazy = FALSE,
                      warning = TRUE,
                      message = TRUE)

## plan("multiprocess", workers = NTHREADS)
## options(future.globals.maxSize= 1.4e9)

## options(bitmapType='cairo')
```

# Aim


> Compare to published datasets (see block1, 2 and 3). The highest priority,m is to compare to those in block 1 (from Abud et al.).

Methods: We processed our RNA-seq, McQuade's TREM raw reads with ARMOR (PMID 31088905). Briefly, reads were aligned and counted to the human genome (GRCh38 assembly and Gencode release 43) with salmon v1.4.0 and with STAR 2.7.7a. Differential expression analysis was carried out with edgeR v3.36.0 using salmon outputs. 

We downloaded Abud's and McQuade's iPSCs data from the Sequence Read Archive (SRA) accessions SRP092075 and SRP155574, respectively, using recount3 (PMID 34844637) and GRCh38 as reference genome and Gencode's annotation. 

Results: We detect 'strong' age effects (within Villani's WTs), 'weak' mutant effects (within Villani's mutant), generalized batch effects which are hard to deal with when comparing to Abud's, McQuade's data. The MDS plot after batch correction of Villani's WT + Villani's mutants + Abud's looks the best; the same MDS without Villani's mutants not so much.

See `Comparison of Villani's, Abud's, McQuade's data`.

> Do they express typical microglial genes (I have a list)

Methods: We generated multi-dimensional scaling (MDS) plots based on normalized count data using edgeR v3.36.0. Briefly, MDS plots depict similarities between samples (including replicates and batches) in an unsupervised manner, depicting the two leading fold-change dimensions which explain the largest proportion of variation in gene expression across samples.

Results: Please use the iSEE deployments:

-
-
-

Mind there you can select 'counts', 'CPMs' or 'logCPMs'. My recommendation would be to check 'CPMs' or 'logCPMs', so you have readouts that don't depend on sequencing depth.


> They express genes relevant in neurodegenerative disorders (I have a list)

Results: See above.


> 1. The different mutants still becomes microglia (add to the MDS?)
>        a. Slc37a2 I have 2 different mutants (clones) same gene (but different mutation)
>        b. TREM2 I just have 1 mutant

>2. They are indeed mutants 
>        a. RNA level reduced compared to WT. As WT for comparison use the 2 from av165 (same round of differentiation)
>        b. Alignment to WT sequence?

Results: please browse the STAR BAM files.

> 3. The TREM2 mutant has the typical signature of microglia missing this gene according to literature
>        a. Compare to McQuade TREM2 KO (highlighted in yellow in the spreadsheet)
>        b. Lipid metabolism affected (expected)


Methods: We ran a geneset enrichment analysis using `camera` from limma v3.50.3 and [mSigDB](http://software.broadinstitute.org/gsea/msigdb) annotations on differentially expressed genes. Selected genesets included:

<!-- - H, hallmark gene sets  -->
<!-- - C1, positional gene sets (cytobands) -->
- C2, curated gene sets from online pathway databases, publications in PubMed, and knowledge of domain experts
<!-- - C3, regulatory target gene sets based on gene target predictions for microRNA seed sequences and predicted transcription factor binding sites -->
- C5, Gene Ontology
- C8, cell type signatures

Results: please check sections named `Geneset analysis`.

# Comparison of Villani's, Abud's, McQuade's data

## Metadata

```{r}
options(ggplot2.discrete.fill    = function() scale_fill_brewer(palette = "Set3"))
options(ggplot2.discrete.color    = function() scale_color_brewer(palette = "Set3"))
```

```{r}
meta <- read.csv(text="treatment,names,age
iPSC_MG_(Abud),SRR4450428,unknown
iPSC_MG_(Abud),SRR4450429,unknown
iPSC_MG_(Abud),SRR4450430,unknown
iPSC_MG_(Abud),SRR4450431,unknown
iPSC_MG_(Abud),SRR4450432,unknown
iPSC_MG_(Abud),SRR4450433,unknown
iHPC_(Abud),SRR4450434,unknown
iHPC_(Abud),SRR4450435,unknown
iHPC_(Abud),SRR4450436,unknown
iPSC_(Abud),SRR4450437,unknown
iPSC_(Abud),SRR4450438,unknown
iPSC_(Abud),SRR4450439,unknown
iPSC_(Abud),SRR4450440,unknown
CD14_M_(Abud),SRR4450441,unknown
CD14_M_(Abud),SRR4450442,unknown
CD14_M_(Abud),SRR4450443,unknown
CD14_M_(Abud),SRR4450444,unknown
CD14_M_(Abud),SRR4450445,unknown
CD16_M_(Abud),SRR4450446,unknown
CD16_M_(Abud),SRR4450447,unknown
CD16_M_(Abud),SRR4450448,unknown
CD16_M_(Abud),SRR4450449,unknown
Fetal_MG_(Abud),SRR4450450,unknown
Fetal_MG_(Abud),SRR4450451,unknown
Fetal_MG_(Abud),SRR4450452,unknown
Adult_MG_(Abud),SRR4450453,unknown
Adult_MG_(Abud),SRR4450454,unknown
Adult_MG_(Abud),SRR4450455,unknown
Blood_DC_(Abud),SRR4450456,unknown
Blood_DC_(Abud),SRR4450457,unknown
Blood_DC_(Abud),SRR4450458,unknown
iPSC_MG_WT_(McQuade),SRR12608405,unknown
iPSC_MG_WT_(McQuade),SRR12608406,unknown
iPSC_MG_WT_(McQuade),SRR12608407,unknown
iPSC_MG_WT_(McQuade),SRR12608408,unknown
iPSC_MG_TREM2_KO_(McQuade),SRR12608409,unknown
iPSC_MG_TREM2_KO_(McQuade),SRR12608410,unknown
iPSC_MG_TREM2_KO_(McQuade),SRR12608411,unknown
iPSC_MG_TREM2_KO_(McQuade),SRR12608412,unknown
iPSC_MG_2.0_(McQuade),SRR7613924,unknown
iPSC_MG_2.0_(McQuade),SRR7613925,unknown
iPSC_MG_2.0_(McQuade),SRR7613926,unknown
iHPC_2.0_(McQuade),SRR8180288,unknown
iHPC_2.0_(McQuade),SRR8180289,unknown
iHPC_2.0_(McQuade),SRR8180290,unknown
iPSC_MG,20220209.A-av111_WT_001_R1,day_34
iPSC_MG,20220209.A-av111_WT_002_R1,day_34
iPSC_MG,20220209.A-av111_WT_003_R1,day_34
iPSC_MG,20220209.A-tw029_WT_2_R1,day_38
iPSC_MG,20220209.A-tw029_WT_3_R1,day_38
iPSC_MG,20220504.B-WTC_1_R1,day_28
iPSC_MG,20220504.B-WTC_2_R1,day_28
iPSC_MG_Slc37a2_mut1,20220504.B-SLC-EX2B1_1_R1,day_28
iPSC_MG_Slc37a2_mut1,20220504.B-SLC-EX2B1_2_R1,day_28
iPSC_MG_Slc37a2_mut2,20220504.B-SLC-EX6C5_1_R1,day_28
iPSC_MG_Slc37a2_mut2,20220504.B-SLC-EX6C5_2_R1,day_28
iPSC_MG_TREM2_mut,20220504.B-TREM2-A1_1_R1,day_28
iPSC_MG_TREM2_mut,20220504.B-TREM2-A1_2_R1,day_28")
```

```{r}
meta$origin <- 'Villani'

meta$origin[grepl('Abud', meta$treatment)] <- 'Abud'
meta$origin[grepl('McQuade', meta$treatment)] <- 'McQuade iPSC'
meta$origin[grepl('SRR126', meta$names)] <- 'McQuade TREM'

```

Data included in this analysis:

```{r, results = 'asis'}
(DT::datatable(meta, rownames = FALSE, 
                extensions = 'Buttons',
                filter = "none",
                options = list(pageLength = 5, autowidth = TRUE,
                           dom = 'Blftip',                           
                           buttons = c('copy', 'csv', 'excel'))))
```


## Overview of inhouse data alone {.tabset .tabset-pills}

```{r}
d <- readRDS('/home/imallona/avillani_microglia/ambra_only_output/outputR/shiny_sce.rds')$sce_gene

```

```{r}

logcpms <- assay(d, "logcpm")
mds <- limma::plotMDS(logcpms, top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE,
                      var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(logcpms),
                      MDS1 = mds$x,
                      MDS2 = mds$y,
                      var.explained = mds$var.explained)
}
mds <- mds %>% 
    dplyr::full_join(data.frame(colData(d)), by = "names")
```

### MDS {.tabset .tabset-pills}


Multidimensional scaling (MDS) plots samples on a two-dimensional scatterplot so that distances on the plot approximate the typical log2 fold changes between the samples.
     
```{r, fig.width = 10, fig.height = 4, results = 'asis'}

for (item in c('treatment', 'timepoint', 'names', 'timepoint')) {
    cat('#### ', item, '\n\n')
    print(ggplot(mds, aes(x=MDS1, y=MDS2, color = get(item)), alpha=1) +
          geom_point(size=3) +
          labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100),
               title= paste("MDS", item)) +
          coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                          ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
          theme_bw() +
          labs(color = item) +
          theme(aspect.ratio=1) +
          theme(legend.position='right')) #, legend.box='vertical', legend.text=element_text(size=9), legend.title=element_text(size=11),
                #axis.text=element_text(size=10), axis.title=element_text(size=12), title=element_text(size=12)))
    cat('\n\n')

}
```

<!-- ### Correlation  {.tabset .tabset-pills} -->

<!-- #### All genes -->

<!-- ```{r, fig.width = 10, fig.height = 10} -->
<!-- corrplot(cor(logcpms, method= 'spearman'), -->
<!--          method  =  'circle', type  =  'lower', insig = 'blank', -->
<!--           number.cex  =  0.8, order  =  'hclust', diag = TRUE , addCoef.col = 'white') -->


<!-- ``` -->

<!-- #### Top 500 most variable genes -->

<!-- ```{r, fig.width = 10, fig.height = 10} -->
<!-- corrplot(cor(logcpms[head(order(rowVars(logcpms), decreasing = TRUE), 500),], method= 'spearman'), -->
<!--          method  =  'circle', type  =  'lower', insig = 'blank', -->
<!--          addCoef.col  = 'white', number.cex  =  0.8, order  =  'hclust', diag = TRUE) -->




<!-- ``` -->


## Villani's wildtypes and mutants and Abud's data {.tabset .tabset-pills}

<!-- ### Correlations -->


```{r, message = FALSE, warning = FALSE}
srp <- "SRP092075"
abud <- recount3::create_rse_manual(
                    project = srp,
                    project_home = "data_sources/sra",
                    organism = "human",
                    annotation = "gencode_v29",
                    type = "gene")

assay(abud, "counts") <- transform_counts(abud)


```

```{r, eval = TRUE}
abud <- subset(abud,
               select = colData(abud)$external_id %in%  meta$names)

## abud_annot <- merge(abud_annot, colData(abud), by.x = 'srr', by.y = 'external_id')

```


```{r}
abudlcpm <- edgeR::cpm(assay(abud, 'counts'),
                       log = TRUE, 
                       prior.count = 2)


```

```{r}
## table(rowData(d)$gene_id %in% rowData(abud)$gene_id)

rowData(d)$unversioned_gene_id <- sapply(strsplit(rowData(d)$gene_id, '\\.'), function(x) return(x[[1]]))
rowData(abud)$unversioned_gene_id <- sapply(strsplit(rowData(abud)$gene_id, '\\.'), function(x) return(x[[1]]))

## table(rowData(d)$unversioned_gene_id %in% rowData(abud)$unversioned_gene_id)

shared <- intersect(rowData(d)$unversioned_gene_id, rowData(abud)$unversioned_gene_id)

dict <- data.frame(unversioned_gene_id = shared,
                   armor = rownames(rowData(d)[rowData(d)$unversioned_gene_id %in% shared,]))

dict <- merge(dict, rowData(abud)[rowData(abud)$unversioned_gene_id %in% shared,
                                  c('unversioned_gene_id', 'gene_id')],
              by = 'unversioned_gene_id')
colnames(dict)[3] <- 'recount'
```


<!-- ```{r, fig.width = 25, fig.height = 25} -->

<!-- ## dim(cbind(logcpms[dict$armor,], abudlcpm[dict$recount,])) -->

<!-- corrplot(cor(cbind(logcpms[dict$armor,], -->
<!--                    abudlcpm[dict$recount,]), method= 'spearman'), -->
<!--          method  =  'circle', type  =  'lower', insig = 'blank', -->
<!--           number.cex  =  0.8, order  =  'hclust', diag = TRUE , addCoef.col = 'white') -->
<!-- ``` -->

### MDS before batch correction

```{r, fig.width = 8, fig.height = 5}

mds <- limma::plotMDS(cbind(logcpms[dict$armor,],
                            abudlcpm[dict$recount,]), top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(cbind(logcpms[dict$armor,],
                            abudlcpm[dict$recount,])),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

## mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'Abud', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```

```{r, fig.width = 10, fig.height = 4, results = 'asis'}

print(ggplot(mds, aes(x=MDS1, y=MDS2, color = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS before ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(aspect.ratio=1) +
      theme(legend.position='right')) #, legend.box='vertical', legend.text=element_text(size=9), legend.title=element_text(size=11),
                                        #axis.text=element_text(size=10), axis.title=element_text(size=12), title=element_text(size=12)))



```

### Batch correction


```{r}
# parametric adjustment
combat_abud <- sva::ComBat(dat=cbind(logcpms[dict$armor,],
                                    abudlcpm[dict$recount,]),
                          batch = mds$origin,
                          mod = NULL, par.prior = TRUE, prior.plots = TRUE)

```

### MDS after batch correction {.tabset .tabset-pills}

#### Batches


```{r, fig.width = 8, fig.height = 5}
mds <- limma::plotMDS(combat_abud, top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(cbind(logcpms[dict$armor,],
                            abudlcpm[dict$recount,])),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

## mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'Abud', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(aspect.ratio=1) +
      theme(legend.position='right')) #, legend.box='vertical', legend.text=element_text(size=9), legend.title=element_text(size=11),
                                        #axis.text=element_text(size=10), axis.title=element_text(size=12), title=element_text(size=12)))
```

#### Annotation

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = treatment, shape = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(legend.position='right') +
      guides(fill=guide_legend(ncol=2)) +
      theme(aspect.ratio=1) +
      scale_color_brewer(palette = "Set3"))

```



## Villani's wildtypes (no mutants) and Abud's data {.tabset .tabset-pills}

```{r}
ambra_wts <- meta[meta$treatment == 'iPSC_MG', 'names']
```


<!-- ```{r, fig.width = 25, fig.height = 25} -->

<!-- ## dim(cbind(logcpms[dict$armor,], abudlcpm[dict$recount,])) -->

<!-- corrplot(cor(cbind(logcpms[dict$armor, ambra_wts], -->
<!--                    abudlcpm[dict$recount,]), method= 'spearman'), -->
<!--          method  =  'circle', type  =  'lower', insig = 'blank', -->
<!--           number.cex  =  0.8, order  =  'hclust', diag = TRUE , addCoef.col = 'white') -->
<!-- ``` -->

### MDS before batch correction {.tabset .tabset-pills}

```{r fids, fig.width = 8, fig.height = 5}

mds <- limma::plotMDS(cbind(logcpms[dict$armor,ambra_wts],
                            abudlcpm[dict$recount,]), top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(cbind(logcpms[dict$armor,ambra_wts],
                            abudlcpm[dict$recount,])),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

## mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'Abud', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```

```{r, fig.width = 10, fig.height = 4, results = 'asis'}

for (item in c('treatment', 'age', 'names')) {
    cat('#### ', item, '\n\n')
    
    print(ggplot(mds, aes(x=MDS1, y=MDS2, color = get(item)), alpha=1) +
          geom_point(size=3) +
         labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS before ComBat")) +
          coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                          ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
          theme_bw() +
          labs(color = item) +
          theme(aspect.ratio=1) +
          theme(legend.position='right')) #, legend.box='vertical', legend.text=element_text(size=9), legend.title=element_text(size=11),
                                        #axis.text=element_text(size=10), axis.title=element_text(size=12), title=element_text(size=12)))
    cat('\n\n')
}


```

### Batch correction


```{r}
# parametric adjustment; explore the parametric too
combat_abud <- sva::ComBat(dat=cbind(logcpms[dict$armor,ambra_wts],
                                    abudlcpm[dict$recount,]),
                          batch = mds$origin,
                          mod = NULL, par.prior = TRUE, prior.plots = FALSE)

```

### MDS after batch correction {.tabset .tabset-pills}


```{r, fig.width = 8, fig.height = 5}
mds <- limma::plotMDS(combat_abud, top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(cbind(logcpms[dict$armor,ambra_wts],
                            abudlcpm[dict$recount,])),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

## mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'Abud', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```

```{r, fig.width = 8, fig.height = 4, results = 'asis'}

for (item in c('treatment', 'names', 'age')) {
    cat('#### ', item, '\n\n')
    
    print(ggplot(mds, aes(x=MDS1, y=MDS2, color = get(item)), alpha=1) +
          geom_point(size=3) +
         labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
          coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                          ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
          theme_bw() +
          theme(aspect.ratio=1) +
          theme(legend.position='right')) #, legend.box='vertical', legend.text=element_text(size=9), legend.title=element_text(size=11),
                                        #axis.text=element_text(size=10), axis.title=element_text(size=12), title=element_text(size=12)))
    cat('\n\n')
}
```

#### Annotation

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = treatment, shape = age), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(legend.position='right') +
      guides(fill=guide_legend(ncol=2)) +
      theme(aspect.ratio=1) +
      scale_color_brewer(palette = "Set3"))

```




## Villani's wildtypes and mutants and McQuade TREM data {.tabset .tabset-pills}

<!-- ```{r} -->

<!-- srp <- 'SRP281230' -->

<!-- trem <- recount3::create_rse_manual( -->
<!--                     project = srp, -->
<!--                     project_home = "data_sources/sra", -->
<!--                     organism = "human", -->
<!--                     annotation = "gencode_v29", -->
<!--                     type = "gene") -->

<!-- trem <- subset(trem, -->
<!--                select = colData(trem)$external_id %in%  meta$names) -->

<!-- assay(trem, "counts") <- transform_counts(trem) -->


<!-- ``` -->


```{r}
trem <- readRDS('/home/imallona/avillani_microglia/mcquade_trem_only_output/outputR/shiny_sce.rds')$sce_gene

trem <- subset(trem,
               select = colData(trem)$names %in%  meta$names)

tremlcpm <- assay(trem, "logcpm")

```




<!-- ### Correlations -->

<!-- ```{r, fig.width = 25, fig.height = 25} -->


<!-- corrplot(cor(cbind(logcpms[dict$armor,], -->
<!--                    tremlcpm[dict$armor,]), method= 'spearman'), -->
<!--          method  =  'circle', type  =  'lower', insig = 'blank', -->
<!--           number.cex  =  0.8, order  =  'hclust', diag = TRUE , addCoef.col = 'white') -->
<!-- ``` -->


### MDS before batch correction





```{r, fig.width = 8, fig.height = 5}

mds <- limma::plotMDS(cbind(logcpms[dict$armor,],
                           tremlcpm[dict$armor,]), top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(cbind(logcpms[dict$armor,],
                            tremlcpm[dict$armor,])),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

## mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'McQuade TREM', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```

```{r, fig.width = 10, fig.height = 4, results = 'asis'}

print(ggplot(mds, aes(x=MDS1, y=MDS2, color = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS before ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(aspect.ratio=1) +
      theme(legend.position='right')) #, legend.box='vertical', legend.text=element_text(size=9), legend.title=element_text(size=11),
                                        #axis.text=element_text(size=10), axis.title=element_text(size=12), title=element_text(size=12)))



```

### Batch correction


```{r}
# parametric adjustment
combat_trem <- sva::ComBat(dat=cbind(logcpms[dict$armor,],
                                    tremlcpm[dict$armor,]),
                          batch = mds$origin,
                          mod = NULL, par.prior = TRUE, prior.plots = TRUE)

```

### MDS after batch correction {.tabset .tabset-pills}

#### Batch


```{r, fig.width = 8, fig.height = 5}
mds <- limma::plotMDS(combat_trem, top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(cbind(logcpms[dict$armor,],
                            tremlcpm[dict$armor,])),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

# mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'Trem', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```



```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(aspect.ratio=1) +
      theme(legend.position='right')) #, legend.box='vertical', legend.text=element_text(size=9), legend.title=element_text(size=11),
                                        #axis.text=element_text(size=10), axis.title=element_text(size=12), title=element_text(size=12)))
```

#### Annotation

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = treatment, shape = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(legend.position='right') +
      guides(fill=guide_legend(ncol=2)) +
      theme(aspect.ratio=1) +
      scale_color_brewer(palette = "Set3"))

```



## Villani's wildtypes and mutants and McQuade's iPSCs data {.tabset .tabset-pills}

```{r, message = FALSE, warning = FALSE}
srp <- 'SRP155574'

mc <- recount3::create_rse_manual(
                    project = srp,
                    project_home = "data_sources/sra",
                    organism = "human",
                    annotation = "gencode_v29",
                    type = "gene")

assay(mc, "counts") <- transform_counts(mc)

```

```{r}
mc <- subset(mc,
               select = colData(mc)$external_id %in%  meta$names)

assay(mc, "counts") <- transform_counts(mc)
```




```{r}
mclcpm <- edgeR::cpm(assay(mc, 'counts'),
                       log = TRUE, 
                       prior.count = 2)


```

```{r}
## table(rowData(d)$gene_id %in% rowData(mc)$gene_id)

rowData(d)$unversioned_gene_id <- sapply(strsplit(rowData(d)$gene_id, '\\.'), function(x) return(x[[1]]))
rowData(mc)$unversioned_gene_id <- sapply(strsplit(rowData(mc)$gene_id, '\\.'), function(x) return(x[[1]]))

## table(rowData(d)$unversioned_gene_id %in% rowData(mc)$unversioned_gene_id)

shared <- intersect(rowData(d)$unversioned_gene_id, rowData(mc)$unversioned_gene_id)

dict <- data.frame(unversioned_gene_id = shared,
                   armor = rownames(rowData(d)[rowData(d)$unversioned_gene_id %in% shared,]))

dict <- merge(dict, rowData(mc)[rowData(mc)$unversioned_gene_id %in% shared, c('unversioned_gene_id', 'gene_id')],
              by = 'unversioned_gene_id')
colnames(dict)[3] <- 'recount'
```

<!-- ### Correlations -->


<!-- ```{r, fig.width = 25, fig.height = 25} -->

<!-- ## dim(cbind(logcpms[dict$armor,], mclcpm[dict$recount,])) -->

<!-- corrplot(cor(cbind(logcpms[dict$armor,], -->
<!--                    mclcpm[dict$recount,]), method= 'spearman'), -->
<!--          method  =  'circle', type  =  'lower', insig = 'blank', -->
<!--           number.cex  =  0.8, order  =  'hclust', diag = TRUE , addCoef.col = 'white') -->
<!-- ``` -->

### MDS before batch correction

```{r, fig.width = 8, fig.height = 5}

mds <- limma::plotMDS(cbind(logcpms[dict$armor,],
                            mclcpm[dict$recount,]), top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(cbind(logcpms[dict$armor,],
                            mclcpm[dict$recount,])),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

#mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'Mc', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```

```{r, fig.width = 10, fig.height = 4, results = 'asis'}

print(ggplot(mds, aes(x=MDS1, y=MDS2, color = origin), alpha=0.5) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS before ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(aspect.ratio = 1) +
      theme(legend.position='right'))



```

### Batch correction


```{r}
# nonparametric adjustment
combat_mc <- sva::ComBat(dat=cbind(logcpms[dict$armor,],
                                    mclcpm[dict$recount,]),
                          batch = mds$origin,
                          mod = NULL, par.prior = TRUE, prior.plots = TRUE)

```

### MDS after batch correction {.tabset .tabset-pills}

#### Batch


```{r, fig.width = 8, fig.height = 5}
mds <- limma::plotMDS(combat_mc, top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(cbind(logcpms[dict$armor,],
                            mclcpm[dict$recount,])),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

#mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'Mc', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(aspect.ratio=1) +
      theme(legend.position='right')) #, legend.box='vertical', legend.text=element_text(size=9), legend.title=element_text(size=11),
                                        #axis.text=element_text(size=10), axis.title=element_text(size=12), title=element_text(size=12)))
```

#### Annotation

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = treatment, shape = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(legend.position='right') +
      guides(fill=guide_legend(ncol=2)) +
      theme(aspect.ratio=1) +
      scale_color_brewer(palette = "Set3"))

```


# Joint embedding of Villani's plus public data, combined {.tabset .tabset-pills}

```{r}
pre <- cbind(logcpms[dict$armor,],
             abudlcpm[dict$recount,],
             tremlcpm[dict$armor,],
             mclcpm[dict$recount,])
```

<!-- ## Correlations (all genes) -->

```{r, fig.width = 30, fig.height = 30}

## dim(cbind(logcpms[dict$armor,], mclcpm[dict$recount,]))

cnames <- colnames(pre)
colnames(pre) <- meta[meta$names == colnames(pre), 'treatment']

## corrplot(cor(pre, method= 'spearman'),
##          method  =  'circle', type  =  'lower', insig = 'blank',
##          number.cex  =  0.8, order  =  'hclust', diag = TRUE , addCoef.col = 'white')


```

<!-- ## Correlations (top 500 variable genes) -->


<!-- ```{r, fig.width = 30, fig.height = 30} -->
<!-- corrplot(cor(pre[head(order(rowVars(pre), decreasing = TRUE), 500),], method= 'spearman'), -->
<!--          method  =  'circle', type  =  'lower', insig = 'blank', -->
<!--          addCoef.col  = 'white', number.cex  =  0.8, order  =  'hclust', diag = TRUE) -->


<!-- ``` -->


```{r}
## take the original colnames back
colnames(pre) <- cnames
```

## MDS before batch correction {.tabset .tabset-pills}

### Batch

```{r firstmds, fig.width = 8, fig.height = 5}

mds <- limma::plotMDS(pre, top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(logcpms) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(pre),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

## mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'public', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```

```{r, fig.width = 10, fig.height = 4, results = 'asis'}

print(ggplot(mds, aes(x=MDS1, y=MDS2, color = origin), alpha=1) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS before ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(aspect.ratio = 1) + 
      theme(legend.position='right')) 


```


## Annotation

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = treatment, shape = origin), alpha=0.5) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(legend.position='right') +
      guides(color=guide_legend(ncol=2)) +
      theme(aspect.ratio=1))

```


## Batch correction


```{r}

pre <- pre[,meta$names]

stopifnot(all(colnames(pre) == meta$names))


# nonparametric adjustment is as bad
post <- sva::ComBat(dat=pre,
                    batch = meta$origin,
                    mod = NULL, par.prior = TRUE, prior.plots = TRUE)

```

## MDS after batch correction {.tabset .tabset-pills}

### Batch


```{r, fig.width = 8, fig.height = 5}
mds <- limma::plotMDS(post, top = 500, labels = NULL, pch = NULL,
                      cex = 1, dim.plot = c(1, 2), ndim = min(7, ncol(post) - 1),
                      gene.selection = "common",
                      xlab = NULL, ylab = NULL, plot = FALSE, var.explained = TRUE)
if (!is.null(mds$cmdscale.out)) {
    ## Bioc 3.12 and earlier
    mds <- mds$cmdscale.out
    colnames(mds) <- paste0("MDS", seq_len(ncol(mds)))
    mds <- as.data.frame(mds) %>% tibble::rownames_to_column(var = "names")
} else {
    mds <- data.frame(names = colnames(post),
                      MDS1 = mds$x,
                      MDS2 = mds$y, var.explained = mds$var.explained)
}

# mds$origin <- ifelse(grepl('SRR', mds$names), yes = 'Mc', no = 'Ambra')

mds <- na.omit( mds %>% dplyr::full_join(data.frame(meta), by = "names"))

```

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = origin), alpha=0.5) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(aspect.ratio=1) +
      theme(legend.position='right'))
```

### Annotation

```{r, fig.width = 8, fig.height = 4, results = 'asis'}
print(ggplot(mds, aes(x=MDS1, y=MDS2, color = treatment, shape = origin), alpha=0.5) +
      geom_point(size=3) +
     labs(x = sprintf("Dim. 1, var. exp. %s%%", round(mds$var.explained[1],2)*100),
               y = sprintf("Dim. 2, var. exp. %s%%", round(mds$var.explained[2], 2)*100), title= paste("MDS after ComBat")) +
      coord_cartesian(xlim=range(mds$MDS1, na.rm=TRUE),
                      ylim=range(mds$MDS2, na.rm=TRUE), expand=TRUE) +
      theme_bw() +
      theme(legend.position='right') +
      guides(color=guide_legend(ncol=2)) +
      theme(aspect.ratio=1))

```

# Differential expression: mutants vs WTs (regardless of the WTs age)

Original ARMOR run, duplicated here to consolidate outputs.

Here, we perform differential gene expression analysis with edgeR 
[@Robinson2010edgeR] followed by gene set analysis with camera [@Wu2012camera],
based on abundance estimates from Salmon. For more detailed information of each
step, please refer to the 
[edgeR user guide](https://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf).



```{r edgeR-print-se}
## List of SummarizedExperiment objects (gene/transcript level)
se <- readRDS('/home/imallona/avillani_microglia/ambra_only_output/outputR/tximeta_se.rds')

## Get gene-level SummarizedExperiment object
sg <- se$sg
metadata <- colData(sg)

sg
```

## Plot total number of reads per sample

```{r edgeR-plot-totalcount}
ggplot(data.frame(totCount = colSums(assay(sg, "counts")), 
                  sample = colnames(assay(sg, "counts")), 
                  stringsAsFactors = FALSE), 
       aes(x = sample, y = totCount)) + geom_bar(stat = "identity") + 
    theme_bw() + xlab("") + ylab("Total read count") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Create DGEList and include average transcript length offsets

A `DGEList` is the main object `edgeR` requires to perform the DGE analysis. It
is designed to store read counts and associated information. After creating this
object, we add offsets, which are average transcript length correction terms 
[@Soneson2016tximport],
and scale them so they are consistent with library sizes (sequencing depth for
each sample).

Then we calculate normalization factors to scale the raw library sizes and
minimize the log-fold changes between the samples for most genes. Here the
trimmed mean of M-values between each pair of samples (TMM) is used by default
[@Robinson2010TMM].

Finally we add gene annotation information.

```{r edgeR-dge-generate}
dge0 <- tximeta::makeDGEList(sg)

dge0$genes <- as.data.frame(rowRanges(sg))
```

## Calculate logCPMs and add as an assay

We calculate log-counts per million (CPMs) because they are useful descriptive
measures for the expression level of a gene. Note, however, that the normalized
values are not used for the differential expression analysis. By default, the
normalized library sizes are used in the computation.

We add the logCPMs to one of the fields (or assay) of the first gene-level
`SummarizedExperiment` object `sg`. At the end of the analysis, we will use this
object again to export the results of all the genes we started with.

```{r edgeR-add-logcpm}
logcpms <- edgeR::cpm(dge0, offset = dge0$offset, log = TRUE, 
                      prior.count = 2)
dimnames(logcpms) <- dimnames(dge0$counts)

stopifnot(all(rownames(logcpms) == rownames(sg)),
          all(colnames(logcpms) == colnames(sg)))
assay(sg, "logcpm") <- logcpms
```

Next, we specify the design matrix of the experiment, defining which sample 
annotations will be taken into account in the statistical modeling.

```{r edgeR-define-design}

design <- '~ 0 + treatment'
metadata <- read.table('/home/imallona/avillani_microglia/ARMOR/metadata_ambra_only.tsv', header = TRUE)
stopifnot(all(colnames(dge0) == metadata$names))


des <- model.matrix(as.formula(design), data = metadata)
```

## Filter out lowly expressed genes

Next we determine which genes have sufficiently large counts to be retained in
the statistical analysis, and remove the rest. After removing genes, we 
recalculate the normalization factors.

```{r edgeR-filter-genes}
dim(dge0)
keep <- edgeR::filterByExpr(dge0, design = des)
dge <- dge0[keep, ]
dim(dge)
```

## Estimate dispersion and fit QL model

We model the count data using a quasi-likelihood (QL) negative binomial (NB)
generalized log-linear model, which accounts for gene-specific variability from
both biological and technical sources. Before fitting the model, we estimate
the NB dispersion (overall biological variability across all genes), and the QL
dispersion (gene-specific) using the `estimateDisp()` function.

It is also good practice to look at the relationship between the biological
coefficient of variation (NB dispersion) and the gene abundance (in logCPMs).

```{r edgeR-estimate-disp}
## Estimate dispersion and fit model
dge <- estimateDisp(dge, design = des)
qlfit <- glmQLFit(dge, design = des)

## Plot dispersions
plotBCV(dge)
```

## Define contrasts 

Before testing for differences in gene expression, we define the contrasts
we wish to test for. Here we represent the constrasts as a numeric matrix:

```{r edgeR-define-contrasts}
contrast <- c('treatmentiPSC_MG_Slc37a2_mut1-treatmentiPSC_MG',
              'treatmentiPSC_MG_Slc37a2_mut2-treatmentiPSC_MG',
              'treatmentiPSC_MG_TREM2_mut-treatmentiPSC_MG',
              'treatmentiPSC_MG_Slc37a2_mut1-treatmentiPSC_MG_TREM2_mut',
              'treatmentiPSC_MG_Slc37a2_mut2-treatmentiPSC_MG_TREM2_mut',
              'treatmentiPSC_MG_Slc37a2_mut2-treatmentiPSC_MG_Slc37a2_mut1')


(contrasts <- as.data.frame(makeContrasts(contrasts = contrast, levels = des)))
```


```{r edgeR-perform-tests}
signif3 <- function(x) signif(x, digits = 3)
edgeR_res <- lapply(contrasts, function(cm) {
    qlf <- glmQLFTest(qlfit, contrast = cm)
    tt <- topTags(qlf, n = Inf, sort.by = "none")$table
    tt %>%
        dplyr::mutate(mlog10PValue = -log10(PValue)) %>% 
        dplyr::mutate_at(vars(one_of(c("logFC", "logCPM", "F", 
                                       "PValue", "FDR", "mlog10PValue"))), 
                         list(signif3))
})
```

## DGE tests and MA plots {.tabset .tabset-pills}

We can visualize the test results by plotting the logCPM (average) vs the logFC,
and coloring genes with an adjusted p-value below 0.05 (or another specificed
FDR threshold). A plot is drawn for every contrast.

```{r, results = 'asis'}
if (is(edgeR_res, "data.frame")) {
    print(ggplot(edgeR_res, aes(x = logCPM, y = logFC, color = FDR <= 0.05)) + 
              geom_point() + theme_bw() + 
              scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")))
} else {
    for (nm in names(edgeR_res)) {
        cat('### ', nm, '\n\n')
        print(ggplot(edgeR_res[[nm]], aes(x = logCPM, y = logFC, color = FDR <= 0.05)) + 
                  geom_point() + theme_bw() + 
                  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) + 
              ggtitle(nm))
        cat('\n\n')
    }
}
```

## Explore DGE results {.tabset .tabset-pills}

We export the results into text files that can be opened using any text editor.

The volcanoes plot the uncorrected p-values, but the coloring (significance) is based on FDR-adjusted p-values.
Independently of the FDRs, genes with an absolute log2 fold-change greater than 0.5 are also reported (coloring and vertical lines will at the negative and positive values of the FC cut-off).


```{r edgeR-save-results, results = 'asis'}
## Write results to text files and make MA plots
if (is(edgeR_res, "data.frame")) {
    write.table(edgeR_res %>% dplyr::arrange(PValue) %>%
                    dplyr::select(-dplyr::any_of("tx_ids")), 
                file = "edgeR_dge_results.txt", 
                sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
} else {
    for (nm in names(edgeR_res)) {
        cat('### ', nm, '\n\n')

        edgeR_res[[nm]]$simple_gene_name <- edgeR_res[[nm]]$gene_name
        edgeR_res[[nm]]$simple_gene_name[is.na(edgeR_res[[nm]]$simple_gene_name)] <- edgeR_res[[nm]]$gene_id[is.na(edgeR_res[[nm]]$simple_gene_name)]
                                                                                                                             
        print(EnhancedVolcano(edgeR_res[[nm]],
                         lab = edgeR_res[[nm]]$simple_gene_name,
                         x = 'logFC',
                         y = 'PValue',
                         pCutoffCol = 'FDR',
                         title = 'differential gene expression',
                         subtitle = nm,
                         pCutoff = 0.05,
                         FCcutoff = 0.5,
                         pointSize = 3.0,
                         drawConnectors = TRUE,
                         labSize = 6.0))
             
        fn <-  paste0("edgeR_dge_results_", nm, "_all_ages.txt")
        write.table(edgeR_res[[nm]] %>% dplyr::arrange(PValue) %>%
                    dplyr::select(-dplyr::any_of("tx_ids")), 
                    file = fn, 
                    sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)

        cat(sprintf('\n\n<a href="%s">Download results</a>\n\n', fn))
        
        ## (DT::datatable(edgeR_res[[nm]], rownames = FALSE, 
        ##                extensions = 'Buttons',
        ##                filter = "none",
        ##                options = list(pageLength = 5, autowidth = TRUE,
        ##                               dom = 'Blftip',                           
        ##                               buttons = c('copy', 'csv', 'excel'))))
        cat('\n\n')
        ## write.table(edgeR_res[[nm]] %>% dplyr::arrange(PValue) %>%
        ##                 dplyr::select(-dplyr::any_of("tx_ids")), 
        ##             file = paste0("edgeR_dge_results_", nm, ".txt"), 
        ##             sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
    }
}
```

<!-- # Output DGE results as list of `SingleCellExperiment` objects -->

<!-- Here, we store the analysis results with the original data. The results are -->
<!-- appended on the `rowData` of the original gene-level `SummarizedExperiment` -->
<!-- object `sg`. For genes that were filtered out, `NA` values are used in the -->
<!-- result columns. The updated `sg` could be fed to the R package `iSEE` to -->
<!-- perform more exploratory and visual analysis. -->

<!-- ```{r edgeR-se} -->
<!-- ## add rows (NA) for genes that are filtered out (if any) -->
<!-- edgeR_resA <- lapply(seq_along(edgeR_res), FUN = function(x) { -->
    
<!--     ## All genes -->
<!--     geneA <- rowData(sg)$gene_id -->
    
<!--     ## Genes that are not filtered out -->
<!--     resX <- edgeR_res[[x]] -->
<!--     resX <- resX %>%  -->
<!--         dplyr::select(c("gene_id", "gene_name", "logFC", "logCPM",  -->
<!--                         "F", "FDR", "PValue", "mlog10PValue"))  -->
<!--     rownames(resX) <- resX$gene_id -->
    
<!--     ## Genes that are filtered out  -->
<!--     geneO <- setdiff(geneA, resX$gene_id) -->
    
<!--     ## results for all genes -->
<!--     if (length(geneO) > 0) { -->
<!--         ## create a data frame with values NA as the results of the genes that -->
<!--         ## are filtered out -->
<!--         matO <- matrix(NA, nrow = length(geneO), -->
<!--                        ncol = ncol(resX), -->
<!--                        dimnames = list(geneO,  -->
<!--                                        colnames(resX)))  -->
<!--         resO <- data.frame(matO) -->
<!--         resO$gene_id <- geneO -->
<!--         resO$gene_name <- rowData(sg)$gene_name[match(geneO, rowData(sg)$gene_id)] -->
        
<!--         ## Combine the result tables  -->
<!--         resA <- resO %>% -->
<!--             dplyr::bind_rows(resX) %>% -->
<!--             dplyr::arrange(match(gene_id, geneA)) %>% -->
<!--             dplyr::mutate(contrast = names(edgeR_res)[[x]])  -->
<!--     } else { -->
<!--         resA <- resX %>% -->
<!--             dplyr::arrange(match(gene_id, geneA)) %>% -->
<!--             dplyr::mutate(contrast = names(edgeR_res)[[x]])  -->
<!--     } -->
    
<!--     ## Use gene column as rownames -->
<!--     rownames(resA) <- paste(resA$gene_id, resA$gene_name, sep = "__") -->
    
<!--     ## convert to DataFrame -->
<!--     resA <- S4Vectors::DataFrame(resA) -->
<!--     return(resA) -->
<!-- }) -->
<!-- names(edgeR_resA) <- names(edgeR_res) -->

<!-- ## Put the result tables in rowData  -->
<!-- for (i in seq_along(edgeR_resA)) { -->
<!--     nam <- names(edgeR_resA)[i] -->
<!--     namI <- paste("edgeR:", nam, sep = "") -->
<!--     stopifnot(all(rownames(sg) == rownames(edgeR_resA[[i]]))) -->
<!--     rowData(sg)[[namI]] <- edgeR_resA[[i]]  -->
<!-- } -->
<!-- ``` -->

<!-- The output is saved as a list. Compared to the input data `se`, the element `sg` -->
<!-- is updated and `st` stays the same. -->

<!-- ```{r edgeR-save-se} -->
<!-- analysis_se <- list(sg = sg, st = se$st) -->
<!-- saveRDS(analysis_se, file = "edgeR_dge.rds") -->
<!-- ``` -->


<!-- ```{r check-gene_names-column, eval = !is.null(genesets), include = FALSE} -->
<!-- if(!("gene_name" %in% colnames(rowData(sg)))) { -->
<!--     genesets <- NULL -->
<!-- } -->
<!-- ``` -->

## Geneset analysis  {.tabset .tabset-pills}


We will use `camera` to perform an enrichment analysis for a collection of
gene sets from the [mSigDB](http://software.broadinstitute.org/gsea/msigdb),
packaged in the `msigdbr` R package. Here, we load the gene set definitions
and select which ones to include in the analysis. Genesets included:

<!-- - H, hallmark gene sets  -->
<!-- - C1, positional gene sets (cytobands) -->
- C2, curated gene sets from online pathway databases, publications in PubMed, and knowledge of domain experts
<!-- - C3, regulatory target gene sets based on gene target predictions for microRNA seed sequences and predicted transcription factor binding sites -->
- C5, Gene Ontology
- C8, cell type signatures




```{r camera-load-genesets, eval = !is.null(genesets), include = !is.null(genesets)}
## genesets <- 'H,C1,C2,C3,C5,C8'
genesets <- 'C2,C5,C8'
genesets <- strsplit(gsub(" ","",genesets), ",")[[1]]

organism <- 'human'

## Retrieve gene sets and combine in a tibble
m_df <- bind_rows(lapply(genesets,
                         function(x) msigdbr(species = organism, category = x)))
```

```{r camera-text2, echo= FALSE, results = 'asis', eval = !is.null(genesets)}
cat("We consider only gene sets where the
number of genes shared with the data set is not too small and not too large.
`camera` is a competitive gene set test that accounts for correlations among
the genes within a gene set.")
```

```{r camera-filter-gene-sets, eval = !is.null(genesets), include = !is.null(genesets)}
minSize <- 3
maxSize <- 500

## Get index for genes in each gene set in the DGEList
indexList <- limma::ids2indices(
    gene.sets = lapply(split(m_df, f = m_df$gs_name), function(w) w$gene_symbol),
    identifiers = dge$genes$gene_name,
    remove.empty = TRUE
)

## Filter out too small or too large gene sets
gsSizes <- vapply(indexList, length, 0)
indexList <- indexList[gsSizes >= minSize & gsSizes <= maxSize]
```

```{r camera-check-indexList-length, eval = !is.null(genesets), include = FALSE}
## Check if the index list is empty after filtering
if (length(indexList) == 0){
  genesets <- NULL
  empty <- TRUE
} else {
  empty <- FALSE
}
```

```{r, echo = FALSE, results = 'asis', eval = !is.null(genesets) && empty}
cat("**NOTE:**
The index list is empty after filtering and `camera` cannot be run. Either try
different gene categories, try different filtering parameters or disable the
gene set analysis in the `config.yaml` file by setting `run_camera: False`.")
```



```{r, eval = !is.null(genesets), include = !is.null(genesets)}
camera_res <- lapply(contrasts, function(cm) {
    camera(dge, index = indexList, design = des, contrast = cm,
           inter.gene.cor = NA)
})
```




```{r camera-save-results, results = 'asis', eval = !is.null(genesets), include = !is.null(genesets)}
## Write results to text files
if (is(camera_res, "data.frame")) {
    ## write.table(camera_res %>% tibble::rownames_to_column("GeneSet") %>%
    ##                 dplyr::arrange(PValue),
    ##             file = "camera_dge_results.txt",
    ##             sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
} else {
    for (nm in names(camera_res)) {
        cat('### ', nm, '\n\n')

        cat('Significantly (before multiple testing correction, pvalue 0.05) enriched/depleted genesets for this contrast:')        
        print(sum(camera_res[[nm]]$PValue < 0.05))
        cat('\n\n')

        
        cat('Significantly (FDR < 0.05) enriched/depleted genesets for this contrast:')        
        print(sum(camera_res[[nm]]$FDR < 0.05))
        cat('\n\n')
        
        fn <- paste0("camera_dge_results_", nm, "_all_ages.txt")
        write.table(camera_res[[nm]] %>%
                        tibble::rownames_to_column("GeneSet") %>%
                        dplyr::arrange(PValue),
                    file = fn,
                    sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)

        cat(sprintf('\n\n<a href="%s">Download geneset analysis results</a>\n\n', fn))
        cat('\n\n')
        
    }
}
```


```{r camera-save-se, eval = !is.null(genesets), include = !is.null(genesets)}
geneSets <- lapply(indexList, function(i) dge$genes$gene_name[i])
saveRDS(list(cameraRes = camera_res,
             geneSets = geneSets), file = "camera_gsa_all_ages.rds")
```



# Differential expression: mutants vs 28-days-old WTs 

## Filter data to remove WTs older than 28 days

```{r}
sg <- sg[,colData(sg)$timepoint == 28]
metadata <- colData(sg)
```

## Plot total number of reads per sample

```{r fig.width = 4, fig.height = 4}
ggplot(data.frame(totCount = colSums(assay(sg, "counts")), 
                  sample = colnames(assay(sg, "counts")), 
                  stringsAsFactors = FALSE), 
       aes(x = sample, y = totCount)) + geom_bar(stat = "identity") + 
    theme_bw() + xlab("") + ylab("Total read count") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

## Create DGEList and include average transcript length offsets

A `DGEList` is the main object `edgeR` requires to perform the DGE analysis. It
is designed to store read counts and associated information. After creating this
object, we add offsets, which are average transcript length correction terms 
[@Soneson2016tximport],
and scale them so they are consistent with library sizes (sequencing depth for
each sample).

Then we calculate normalization factors to scale the raw library sizes and
minimize the log-fold changes between the samples for most genes. Here the
trimmed mean of M-values between each pair of samples (TMM) is used by default
[@Robinson2010TMM].

Finally we add gene annotation information.

```{r}
dge0 <- tximeta::makeDGEList(sg)

dge0$genes <- as.data.frame(rowRanges(sg))
```

## Calculate logCPMs and add as an assay

We calculate log-counts per million (CPMs) because they are useful descriptive
measures for the expression level of a gene. Note, however, that the normalized
values are not used for the differential expression analysis. By default, the
normalized library sizes are used in the computation.

We add the logCPMs to one of the fields (or assay) of the first gene-level
`SummarizedExperiment` object `sg`. At the end of the analysis, we will use this
object again to export the results of all the genes we started with.

```{r}
logcpms <- edgeR::cpm(dge0, offset = dge0$offset, log = TRUE, 
                      prior.count = 2)
dimnames(logcpms) <- dimnames(dge0$counts)

stopifnot(all(rownames(logcpms) == rownames(sg)),
          all(colnames(logcpms) == colnames(sg)))
assay(sg, "logcpm") <- logcpms
```

Next, we specify the design matrix of the experiment, defining which sample 
annotations will be taken into account in the statistical modeling.

```{r}

design <- '~ 0 + treatment'
## metadata <- read.table('/home/imallona/avillani_microglia/ARMOR/metadata_ambra_only.tsv', header = TRUE)
stopifnot(all(colnames(dge0) == metadata$names))


des <- model.matrix(as.formula(design), data = metadata)
```

## Filter out lowly expressed genes

Next we determine which genes have sufficiently large counts to be retained in
the statistical analysis, and remove the rest. After removing genes, we 
recalculate the normalization factors.

```{r}
dim(dge0)
keep <- edgeR::filterByExpr(dge0, design = des)
dge <- dge0[keep, ]
dim(dge)
```

## Estimate dispersion and fit QL model

We model the count data using a quasi-likelihood (QL) negative binomial (NB)
generalized log-linear model, which accounts for gene-specific variability from
both biological and technical sources. Before fitting the model, we estimate
the NB dispersion (overall biological variability across all genes), and the QL
dispersion (gene-specific) using the `estimateDisp()` function.

It is also good practice to look at the relationship between the biological
coefficient of variation (NB dispersion) and the gene abundance (in logCPMs).

```{r}
## Estimate dispersion and fit model
dge <- estimateDisp(dge, design = des)
qlfit <- glmQLFit(dge, design = des)

## Plot dispersions
plotBCV(dge)
```

## Define contrasts 

Before testing for differences in gene expression, we define the contrasts
we wish to test for. Here we represent the constrasts as a numeric matrix:

```{r}
contrast <- c('treatmentiPSC_MG_Slc37a2_mut1-treatmentiPSC_MG',
              'treatmentiPSC_MG_Slc37a2_mut2-treatmentiPSC_MG',
              'treatmentiPSC_MG_TREM2_mut-treatmentiPSC_MG',
              'treatmentiPSC_MG_Slc37a2_mut1-treatmentiPSC_MG_TREM2_mut',
              'treatmentiPSC_MG_Slc37a2_mut2-treatmentiPSC_MG_TREM2_mut',
              'treatmentiPSC_MG_Slc37a2_mut2-treatmentiPSC_MG_Slc37a2_mut1')


(contrasts <- as.data.frame(makeContrasts(contrasts = contrast, levels = des)))
```


```{r}
signif3 <- function(x) signif(x, digits = 3)
edgeR_res <- lapply(contrasts, function(cm) {
    qlf <- glmQLFTest(qlfit, contrast = cm)
    tt <- topTags(qlf, n = Inf, sort.by = "none")$table
    tt %>%
        dplyr::mutate(mlog10PValue = -log10(PValue)) %>% 
        dplyr::mutate_at(vars(one_of(c("logFC", "logCPM", "F", 
                                       "PValue", "FDR", "mlog10PValue"))), 
                         list(signif3))
})
```

## DGE tests and MA plots {.tabset .tabset-pills}

We can visualize the test results by plotting the logCPM (average) vs the logFC,
and coloring genes with an adjusted p-value below 0.05 (or another specificed
FDR threshold). A plot is drawn for every contrast.

```{r, results = 'asis'}
if (is(edgeR_res, "data.frame")) {
    print(ggplot(edgeR_res, aes(x = logCPM, y = logFC, color = FDR <= 0.05)) + 
              geom_point() + theme_bw() + 
              scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")))
} else {
    for (nm in names(edgeR_res)) {
        cat('### ', nm, '\n\n')
        print(ggplot(edgeR_res[[nm]], aes(x = logCPM, y = logFC, color = FDR <= 0.05)) + 
                  geom_point() + theme_bw() + 
                  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) + 
              ggtitle(nm))
        cat('\n\n')
    }
}
```

## Explore DGE results {.tabset .tabset-pills}

We export the results into text files that can be opened using any text editor.

The volcanoes plot the uncorrected p-values, but the coloring (significance) is based on FDR-adjusted p-values.
Independently of the FDRs, genes with an absolute log2 fold-change greater than 0.5 are also reported (coloring and vertical lines will at the negative and positive values of the FC cut-off).


```{r, results = 'asis'}
## Write results to text files and make MA plots
if (is(edgeR_res, "data.frame")) {
    write.table(edgeR_res %>% dplyr::arrange(PValue) %>%
                    dplyr::select(-dplyr::any_of("tx_ids")), 
                file = "edgeR_dge_results.txt", 
                sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
} else {
    for (nm in names(edgeR_res)) {
        cat('### ', nm, '\n\n')

        edgeR_res[[nm]]$simple_gene_name <- edgeR_res[[nm]]$gene_name
        edgeR_res[[nm]]$simple_gene_name[is.na(edgeR_res[[nm]]$simple_gene_name)] <- edgeR_res[[nm]]$gene_id[is.na(edgeR_res[[nm]]$simple_gene_name)]
                                                                                                                             
        print(EnhancedVolcano(edgeR_res[[nm]],
                         lab = edgeR_res[[nm]]$simple_gene_name,
                         x = 'logFC',
                         y = 'PValue',
                         pCutoffCol = 'FDR',
                         title = 'differential gene expression',
                         subtitle = nm,
                         pCutoff = 0.05,
                         FCcutoff = 0.5,
                         drawConnectors = TRUE,
                         pointSize = 3.0,
                         labSize = 6.0))
             
        fn <-  paste0("edgeR_dge_results_", nm, "_28_days_only.txt")
        write.table(edgeR_res[[nm]] %>% dplyr::arrange(PValue) %>%
                    dplyr::select(-dplyr::any_of("tx_ids")), 
                    file = fn, 
                    sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)

        cat(sprintf('\n\n<a href="%s">Download results</a>\n\n', fn))
        
        ## (DT::datatable(edgeR_res[[nm]], rownames = FALSE, 
        ##                extensions = 'Buttons',
        ##                filter = "none",
        ##                options = list(pageLength = 5, autowidth = TRUE,
        ##                               dom = 'Blftip',                           
        ##                               buttons = c('copy', 'csv', 'excel'))))
        cat('\n\n')
        ## write.table(edgeR_res[[nm]] %>% dplyr::arrange(PValue) %>%
        ##                 dplyr::select(-dplyr::any_of("tx_ids")), 
        ##             file = paste0("edgeR_dge_results_", nm, ".txt"), 
        ##             sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
    }
}
```

<!-- # Output DGE results as list of `SingleCellExperiment` objects -->

<!-- Here, we store the analysis results with the original data. The results are -->
<!-- appended on the `rowData` of the original gene-level `SummarizedExperiment` -->
<!-- object `sg`. For genes that were filtered out, `NA` values are used in the -->
<!-- result columns. The updated `sg` could be fed to the R package `iSEE` to -->
<!-- perform more exploratory and visual analysis. -->

<!-- ```{r edgeR-se} -->
<!-- ## add rows (NA) for genes that are filtered out (if any) -->
<!-- edgeR_resA <- lapply(seq_along(edgeR_res), FUN = function(x) { -->
    
<!--     ## All genes -->
<!--     geneA <- rowData(sg)$gene_id -->
    
<!--     ## Genes that are not filtered out -->
<!--     resX <- edgeR_res[[x]] -->
<!--     resX <- resX %>%  -->
<!--         dplyr::select(c("gene_id", "gene_name", "logFC", "logCPM",  -->
<!--                         "F", "FDR", "PValue", "mlog10PValue"))  -->
<!--     rownames(resX) <- resX$gene_id -->
    
<!--     ## Genes that are filtered out  -->
<!--     geneO <- setdiff(geneA, resX$gene_id) -->
    
<!--     ## results for all genes -->
<!--     if (length(geneO) > 0) { -->
<!--         ## create a data frame with values NA as the results of the genes that -->
<!--         ## are filtered out -->
<!--         matO <- matrix(NA, nrow = length(geneO), -->
<!--                        ncol = ncol(resX), -->
<!--                        dimnames = list(geneO,  -->
<!--                                        colnames(resX)))  -->
<!--         resO <- data.frame(matO) -->
<!--         resO$gene_id <- geneO -->
<!--         resO$gene_name <- rowData(sg)$gene_name[match(geneO, rowData(sg)$gene_id)] -->
        
<!--         ## Combine the result tables  -->
<!--         resA <- resO %>% -->
<!--             dplyr::bind_rows(resX) %>% -->
<!--             dplyr::arrange(match(gene_id, geneA)) %>% -->
<!--             dplyr::mutate(contrast = names(edgeR_res)[[x]])  -->
<!--     } else { -->
<!--         resA <- resX %>% -->
<!--             dplyr::arrange(match(gene_id, geneA)) %>% -->
<!--             dplyr::mutate(contrast = names(edgeR_res)[[x]])  -->
<!--     } -->
    
<!--     ## Use gene column as rownames -->
<!--     rownames(resA) <- paste(resA$gene_id, resA$gene_name, sep = "__") -->
    
<!--     ## convert to DataFrame -->
<!--     resA <- S4Vectors::DataFrame(resA) -->
<!--     return(resA) -->
<!-- }) -->
<!-- names(edgeR_resA) <- names(edgeR_res) -->

<!-- ## Put the result tables in rowData  -->
<!-- for (i in seq_along(edgeR_resA)) { -->
<!--     nam <- names(edgeR_resA)[i] -->
<!--     namI <- paste("edgeR:", nam, sep = "") -->
<!--     stopifnot(all(rownames(sg) == rownames(edgeR_resA[[i]]))) -->
<!--     rowData(sg)[[namI]] <- edgeR_resA[[i]]  -->
<!-- } -->
<!-- ``` -->

<!-- The output is saved as a list. Compared to the input data `se`, the element `sg` -->
<!-- is updated and `st` stays the same. -->

<!-- ```{r edgeR-save-se} -->
<!-- analysis_se <- list(sg = sg, st = se$st) -->
<!-- saveRDS(analysis_se, file = "edgeR_dge.rds") -->
<!-- ``` -->


<!-- ```{r check-gene_names-column, eval = !is.null(genesets), include = FALSE} -->
<!-- if(!("gene_name" %in% colnames(rowData(sg)))) { -->
<!--     genesets <- NULL -->
<!-- } -->
<!-- ``` -->


## Geneset analysis {.tabset .tabset-pills}


```{r, eval = !is.null(genesets), include = !is.null(genesets)}
## genesets <- 'H,C1,C2,C3,C5,C8'
genesets <- 'C2,C5,C8'
genesets <- strsplit(gsub(" ","",genesets), ",")[[1]]

organism <- 'human'

## Retrieve gene sets and combine in a tibble
m_df <- bind_rows(lapply(genesets,
                         function(x) msigdbr(species = organism, category = x)))

minSize <- 3
maxSize <- 500

## Get index for genes in each gene set in the DGEList
indexList <- limma::ids2indices(
    gene.sets = lapply(split(m_df, f = m_df$gs_name), function(w) w$gene_symbol),
    identifiers = dge$genes$gene_name,
    remove.empty = TRUE
)

## Filter out too small or too large gene sets
gsSizes <- vapply(indexList, length, 0)
indexList <- indexList[gsSizes >= minSize & gsSizes <= maxSize]
```

```{r, eval = !is.null(genesets), include = FALSE}
## Check if the index list is empty after filtering
if (length(indexList) == 0){
  genesets <- NULL
  empty <- TRUE
} else {
  empty <- FALSE
}
```

```{r, eval = !is.null(genesets), include = !is.null(genesets)}
geneSets <- lapply(indexList, function(i) dge$genes$gene_name[i])
saveRDS(list(cameraRes = camera_res,
             geneSets = geneSets), file = "camera_gsa_28_days_only.rds")
```



```{r camera-perform-tests, cache = FALSE, eval = !is.null(genesets), include = !is.null(genesets)}
camera_res <- lapply(contrasts, function(cm) {
    camera(dge, index = indexList, design = des, contrast = cm,
           inter.gene.cor = NA)
})
```



```{r, results = 'asis', eval = !is.null(genesets), include = !is.null(genesets)}
## Write results to text files
if (is(camera_res, "data.frame")) {
    ## write.table(camera_res %>% tibble::rownames_to_column("GeneSet") %>%
    ##                 dplyr::arrange(PValue),
    ##             file = "camera_dge_results.txt",
    ##             sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)
} else {
    for (nm in names(camera_res)) {
        cat('### ', nm, '\n\n')


        cat('Significantly (before multiple testing correction, pvalue 0.05) enriched/depleted genesets for this contrast:')        
        print(sum(camera_res[[nm]]$PValue < 0.05))
        cat('\n\n')

        
        cat('Significantly (FDR < 0.05) enriched/depleted genesets for this contrast:')        
        print(sum(camera_res[[nm]]$FDR < 0.05))
        cat('\n\n')
        
        fn <- paste0("camera_dge_results_", nm, "_28_days_only.txt")
        write.table(camera_res[[nm]] %>%
                        tibble::rownames_to_column("GeneSet") %>%
                        dplyr::arrange(PValue),
                    file = fn,
                    sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)

        cat(sprintf('\n\n<a href="%s">Download geneset analysis results</a>\n\n', fn))
        cat('\n\n')
        
    }
}
```



# iSEE instances and data objects


```{r, results = 'asis'}

```



# Timestamp

```{r sessionInfo2, cache = FALSE}
date()
sessionInfo()
## devtools::session_info()
```
