---
title: "Ambra Villani's microglia RNA-seq, custom plots"
author: "Izaskun Mallona, Mark D. Robinson's lab, UZH"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: kable
    theme: lumen
params:
  seed: 665
---



```{r warning = FALSE, message = FALSE}

library(ggplot2)
library(SingleCellExperiment)
library(Cairo)
library(scater)
library(iSEE)
library(ComplexHeatmap)

## options(bitmapType="cairo")
```

```{r}
## render on error
knitr::knit_hooks$set(error = function(x, options) {
    knitr::knit_exit()  
})
```

```{r}
knitr::opts_chunk$set(fig.width = 4,
                      fig.height = 5,
                      cache = FALSE,
                      include = TRUE,
                      fig.path = "plots_heatmaps/",
                      dev = c('svg', "png"),
                      cache.lazy = FALSE,
                      warning = TRUE,
                      message = TRUE)

## plan("multiprocess", workers = NTHREADS)
## options(future.globals.maxSize= 1.4e9)

## options(bitmapType='cairo')
```


# Aim

Update of 04 with:

<details>
Nothing major and no re-analysis needed: during review we decided to remove TMEM119 to avoid overinterpretation, so we would simply need to update the RNA-seq plots accordingly.

Concretely, this concerns:

Supplementary Figure 1C and 1D: removal of TMEM119 from the gene lists / plots.

</details>
Custom re-plotting, by Ambra:

<details>

```
I would like the following order of samples:

hiPSC (Abud)
hiHPC (Abud)
Blood DC (Abud)
CD14 M (Abud)
CD16 M (Abud)
Etal MG (Abud)
Adult MG (Abud)
hiPSC-MG (Abud)
hiPSC-MG (this work)
 

Could you also plot a second version where you include also the mutants?  Just to have it 😊

The genes I picked, are signature for #3 specific cell types, so some should stay grouped together. And I would like them ordered vertically as listed below (1, 2, 3). The order of the genes in each group doesn’t matter to me, so I follow whatever you suggest (probably grouping/clustering by intensity?). Here are the #3 groups/gene signatures:

pluripotency (hiPSC signature)
POU5F1
NANOG
NODAL
TDGF1
LARP7
ZIC3
VRIN
CNMD
CER1
GDF3.
DPPA4
L1TD1
GAL
TERF1
GRPR
LEFTY2
SEPHS1
PMAIP1
MSH2
BPTF
SCLY
METTL21A
PHC1
HPC signature
CD34
PTPRC
SPN

Microglia signature (all the rest) [in italic I added genes I am not sure the name is correct as I retrieved them from the png extracted from iSEE and they are hidden partially behind the intensity bar]
P2RY12
AIF1
SPI1
TMEM119
TREM2
ITGAM
APOE
MERTK
ITGB5
CX3C31
TGFBP
PROS1
GPR3
CABLE
BHLHE
OLFM
SLCO
SLC7A8
PPARD
CRYBB1
PLA2G15
CTSD
NPL
ACP2
CTSL
P2RX4
ABHD12
MMP9
CCL2
CH3L1
ENG
C1QC
C1QB
C1QA
TMEM86A
HPGDS
ADORA3
RGL1
DAB2
CCL4
STAB1
PMEPA1
IL1A
PALD1
AXL
ETV5
GPR84
P2RY6
ADGRG1
CCR6
LTC4S
RGS10
RUNX1
CSF1R
GAS6
CD33
EGR1
ABCA7
PILRA
ADGRE1
CXCR1
ITGAX
LPL
```

</details>

# Note

To be knitted with `/usr/bin/R`

## Heatmap without mutants

Following Ambra's:


Using her code from iSEE (thanks!!). The sce the version with Abud; md5summed with penticton to make sure they're aligned. They are.


```{r}

se <- readRDS('../03_custom/villani_plus_abud_sce.rds')


## remove mutants
se <- se[,se$treatment %in% c( "iPSC_MG",
            "iPSC_MG_(Abud)",
            "iHPC_(Abud)",         
            "iPSC_(Abud)",
            "CD14_M_(Abud)",
            "CD16_M_(Abud)",
            "Fetal_MG_(Abud)",
            "Adult_MG_(Abud)",
            "Blood_DC_(Abud)")]

## The following list of commands will generate the plots created in iSEE
## Copy them into a script or an R session containing your SingleCellExperiment.
## All commands below refer to your SingleCellExperiment object as `se`.

colormap <- ExperimentColorMap()
colormap <- synchronizeAssays(colormap, se)
all_contents <- list()

################################################################################
## Complex heatmap 1
################################################################################


.chosen.rows <- c("ENSG00000204531.21__POU5F1", "ENSG00000111704.11__NANOG", "ENSG00000156574.10__NODAL", 
    "ENSG00000241186.10__TDGF1", "ENSG00000174720.18__LARP7", "ENSG00000156925.12__ZIC3", 
    "ENSG00000133980.5__VRTN", "ENSG00000136110.13__CNMD", "ENSG00000147869.5__CER1", 
    "ENSG00000184344.4__GDF3", "ENSG00000121570.13__DPPA4", "ENSG00000240563.2__L1TD1", 
    "ENSG00000069482.7__GAL", "ENSG00000147601.15__TERF1", "ENSG00000126010.6__GRPR", 
    "ENSG00000143768.13__LEFTY2", "ENSG00000086475.15__SEPHS1", "ENSG00000141682.12__PMAIP1", 
    "ENSG00000095002.15__MSH2", "ENSG00000171634.19__BPTF", "ENSG00000132330.18__SCLY", 
    "ENSG00000144401.14__METTL21A", "ENSG00000111752.11__PHC1", "ENSG00000174059.17__CD34", 
    "ENSG00000081237.21__PTPRC", "ENSG00000197471.12__SPN", "ENSG00000169313.10__P2RY12", 
    "ENSG00000204472.13__AIF1", "ENSG00000066336.12__SPI1",
    "ENSG00000095970.17__TREM2", "ENSG00000169896.18__ITGAM", "ENSG00000130203.10__APOE", 
    "ENSG00000153208.18__MERTK", "ENSG00000082781.12__ITGB5", "ENSG00000168329.14__CX3CR1", 
    "ENSG00000106799.14__TGFBR1", "ENSG00000184500.16__PROS1", "ENSG00000171659.16__GPR34", 
    "ENSG00000134508.13__CABLES1", "ENSG00000123095.6__BHLHE41", "ENSG00000116774.12__OLFML3", 
    "ENSG00000137491.15__SLCO2B1", "ENSG00000092068.21__SLC7A8", "ENSG00000112033.14__PPARD", 
    "ENSG00000100122.7__CRYBB1", "ENSG00000103066.13__PLA2G15", "ENSG00000117984.15__CTSD", 
    "ENSG00000135838.14__NPL", "ENSG00000134575.13__ACP2", "ENSG00000135047.16__CTSL", 
    "ENSG00000135124.16__P2RX4", "ENSG00000100997.20__ABHD12", "ENSG00000100985.7__MMP9", 
    "ENSG00000108691.10__CCL2", "ENSG00000133048.13__CHI3L1", "ENSG00000106991.14__ENG", 
    "ENSG00000159189.13__C1QC", "ENSG00000173369.18__C1QB", "ENSG00000173372.18__C1QA", 
    "ENSG00000151117.9__TMEM86A", "ENSG00000163106.11__HPGDS", "ENSG00000282608.2__ADORA3", 
    "ENSG00000143344.16__RGL1", "ENSG00000153071.15__DAB2", "ENSG00000275302.2__CCL4", 
    "ENSG00000010327.10__STAB1", "ENSG00000124225.16__PMEPA1", "ENSG00000115008.6__IL1A", 
    "ENSG00000107719.10__PALD1", "ENSG00000167601.12__AXL", "ENSG00000244405.8__ETV5", 
    "ENSG00000139572.4__GPR84", "ENSG00000171631.16__P2RY6", "ENSG00000112486.18__CCR6", 
    "ENSG00000205336.14__ADGRG1", "ENSG00000213316.10__LTC4S", "ENSG00000148908.15__RGS10", 
    "ENSG00000159216.19__RUNX1", "ENSG00000182578.14__CSF1R", "ENSG00000183087.15__GAS6", 
    "ENSG00000105383.15__CD33", "ENSG00000120738.8__EGR1", "ENSG00000064687.13__ABCA7", 
    "ENSG00000085514.16__PILRA", "ENSG00000174837.15__ADGRE1", "ENSG00000163464.8__CXCR1", 
    "ENSG00000140678.17__ITGAX", "ENSG00000175445.17__LPL");

```


```{r}
## catalog genes
pluri <- read.table(text="POU5F1
NANOG
NODAL
TDGF1
LARP7
ZIC3
VRTN
CNMD
CER1
GDF3
DPPA4
L1TD1
GAL
TERF1
GRPR
LEFTY2
SEPHS1
PMAIP1
MSH2
BPTF
SCLY
METTL21A
PHC1")$V1

hpc <- read.table(text="CD34
PTPRC
SPN")$V1

micro <- read.table(text="P2RY12
AIF1
SPI1
# TMEM119
TREM2
ITGAM
APOE
MERTK
ITGB5
CX3CR1
TGFBR1
PROS1
GPR34
CABLES1
BHLHE41
OLFML3
SLCO2B1
SLC7A8
PPARD
CRYBB1
PLA2G15
CTSD
NPL
ACP2
CTSL
P2RX4
ABHD12
MMP9
CCL2
CHI3L1
ENG
C1QC
C1QB
C1QA
TMEM86A
HPGDS
ADORA3
RGL1
DAB2
CCL4
STAB1
PMEPA1
IL1A
PALD1
AXL
ETV5
GPR84
P2RY6
ADGRG1
CCR6
LTC4S
RGS10
RUNX1
CSF1R
GAS6
CD33
EGR1
ABCA7
PILRA
ADGRE1
CXCR1
ITGAX
LPL")$V1

symbols_rows <- sapply(strsplit(.chosen.rows, '__'), function(x) return(x[2]))
stopifnot(all(pluri %in% symbols_rows))
stopifnot(all(hpc %in% symbols_rows))
stopifnot(all(micro %in% symbols_rows))

```

```{r}
## now, sort the columns as requested
##   1.  hiPSC (Abud)
##   2.  hiHPC (Abud)
##   3.  Blood DC (Abud)
##   4.  CD14 M (Abud)
##   5.  CD16 M (Abud)
##   6.  Etal MG (Abud)
##   7.  Adult MG (Abud)
##   8.  hiPSC-MG (Abud)
## 9.  hiPSC-MG (this work)
sorted_cols <- c( "iPSC_(Abud)",  "iHPC_(Abud)", "Blood_DC_(Abud)" ,    
                 "CD14_M_(Abud)",   "CD16_M_(Abud)",
                 "Fetal_MG_(Abud)",  "Adult_MG_(Abud)",
                 "iPSC_MG_(Abud)", "iPSC_MG")

colData(se)$treatment <- factor(colData(se)$treatment, levels = sorted_cols)


```


```{r}
.chosen.columns <- colnames(se);
plot.data <- assay(se, "logcpms")[.chosen.rows, .chosen.columns, drop=FALSE]
plot.data <- as.matrix(plot.data);

plot.data <- plot.data - rowMeans(plot.data)
plot.data <- plot.data / apply(plot.data, 1, sd)

.assay_colors <- c("blue", "white", "red")
.assay_colors <- circlize::colorRamp2(breaks = c(-3.73423928357544, 0, 3.44073249010621), colors = .assay_colors)

# Keep all samples to compute the full range of continuous annotations
.column_data <- colData(se)[, c("origin", "treatment"), drop=FALSE]

.column_col <- list()

.column_data[["origin"]] <- factor(.column_data[["origin"]]);
.color_values <- .column_data[["origin"]]
.color_values <- setdiff(unique(.color_values), NA)
.col_colors <- colDataColorMap(colormap, "origin", discrete=TRUE)(length(.color_values))
names(.col_colors) <- unique(.color_values)
.column_col[["origin"]] <- .col_colors

.column_data[["treatment"]] <- factor(.column_data[["treatment"]]);
.color_values <- .column_data[["treatment"]]
.color_values <- setdiff(unique(.color_values), NA)
.col_colors <- colDataColorMap(colormap, "treatment", discrete=TRUE)(length(.color_values))
names(.col_colors) <- unique(.color_values)
.column_col[["treatment"]] <- .col_colors

.column_data <- .column_data[colnames(plot.data), , drop=FALSE]
.column_data <- as.data.frame(.column_data, optional=TRUE)
.column_annot_order <- order(.column_data[["origin"]], .column_data[["treatment"]])
.column_data <- .column_data[.column_annot_order, , drop=FALSE]
plot.data <- plot.data[, .column_annot_order, drop=FALSE]
.column_annot <- ComplexHeatmap::columnAnnotation(df=.column_data, col=.column_col, annotation_legend_param=list(direction="vertical"))
```

Shortened and sorted gene names:

```{r}
rownames(plot.data) <- sapply(strsplit(rownames(plot.data), '__'), function(x) return(x[2]))
plot.data <- plot.data[c(pluri, hpc, micro),]
row.split <- data.frame(gene = c(pluri, hpc, micro),
                        category = NA)
row.split[row.split$gene %in% pluri, 'category'] <- 'Pluripotency'
row.split[row.split$gene %in% hpc, 'category'] <- 'HPC'
row.split[row.split$gene %in% micro, 'category'] <- 'Microglia'


rowha <- rowAnnotation(signature = row.split$category,
                       show_annotation_name = TRUE)

hm <- ComplexHeatmap::Heatmap(matrix=plot.data, col=.assay_colors,
                              top_annotation=.column_annot, cluster_rows=TRUE,
                              split = as.factor(row.split$category),
                              left_annotation = rowha,
                              clustering_distance_rows="spearman", clustering_method_rows="ward.D2",
                              cluster_columns=FALSE, name="logcpms\n(centered, scaled)",
                              show_row_names=TRUE, show_column_names=TRUE,
                              heatmap_legend_param=list(direction="vertical"))

```


```{r heatmap_nomutants, fig.width = 12, fig.height = 16, results = 'as-is'}
ComplexHeatmap::draw(hm, heatmap_legend_side="right", annotation_legend_side="right")

```

## Heatmap without mutants plotting logCPMs - not scaling nor centering

Because the scaling and centering, even if standard in the field, force heatmap cells to go blue -> red (rowwise). Adding this a a sanity check to be aware of actual logCPMs -hoping it helps for interpretation.



```{r}
.chosen.columns <- colnames(se);
plot.data <- assay(se, "logcpms")[.chosen.rows, .chosen.columns, drop=FALSE]
plot.data <- as.matrix(plot.data);

## no scaling!!
## plot.data <- plot.data - rowMeans(plot.data)
## plot.data <- plot.data / apply(plot.data, 1, sd)

.assay_colors <- c("blue", "white", "red")
## .assay_colors <- circlize::colorRamp2(breaks = c(-3.73423928357544, 0, 3.44073249010621), colors = .assay_colors)

# Keep all samples to compute the full range of continuous annotations
.column_data <- colData(se)[, c("origin", "treatment"), drop=FALSE]

.column_col <- list()

.column_data[["origin"]] <- factor(.column_data[["origin"]]);
.color_values <- .column_data[["origin"]]
.color_values <- setdiff(unique(.color_values), NA)
.col_colors <- colDataColorMap(colormap, "origin", discrete=TRUE)(length(.color_values))
names(.col_colors) <- unique(.color_values)
.column_col[["origin"]] <- .col_colors

.column_data[["treatment"]] <- factor(.column_data[["treatment"]]);
.color_values <- .column_data[["treatment"]]
.color_values <- setdiff(unique(.color_values), NA)
.col_colors <- colDataColorMap(colormap, "treatment", discrete=TRUE)(length(.color_values))
names(.col_colors) <- unique(.color_values)
.column_col[["treatment"]] <- .col_colors

.column_data <- .column_data[colnames(plot.data), , drop=FALSE]
.column_data <- as.data.frame(.column_data, optional=TRUE)
.column_annot_order <- order(.column_data[["origin"]], .column_data[["treatment"]])
.column_data <- .column_data[.column_annot_order, , drop=FALSE]
plot.data <- plot.data[, .column_annot_order, drop=FALSE]
.column_annot <- ComplexHeatmap::columnAnnotation(df=.column_data, col=.column_col, annotation_legend_param=list(direction="vertical"))
```

```{r}

rownames(plot.data) <- sapply(strsplit(rownames(plot.data), '__'), function(x) return(x[2]))
plot.data <- plot.data[c(pluri, hpc, micro),]
row.split <- data.frame(gene = c(pluri, hpc, micro),
                        category = NA)
row.split[row.split$gene %in% pluri, 'category'] <- 'Pluripotency'
row.split[row.split$gene %in% hpc, 'category'] <- 'HPC'
row.split[row.split$gene %in% micro, 'category'] <- 'Microglia'


rowha <- rowAnnotation(signature = row.split$category,
                       show_annotation_name = TRUE)

hm <- ComplexHeatmap::Heatmap(matrix=plot.data, col=.assay_colors,
                              top_annotation=.column_annot, cluster_rows=TRUE,
                              split = as.factor(row.split$category),
                              left_annotation = rowha,
                              clustering_distance_rows="spearman", clustering_method_rows="ward.D2",
                              cluster_columns=FALSE, name="logCPMs",
                              show_row_names=TRUE, show_column_names=TRUE,
                              heatmap_legend_param=list(direction="vertical"))

```


```{r heatmap_nomutants_no_scaling, fig.width = 12, fig.height = 16, results = 'as-is'}
ComplexHeatmap::draw(hm, heatmap_legend_side="right", annotation_legend_side="right")

```



## Heatmap with mutants

With logCPM scaling and centering again.

```{r}
se <- readRDS('../03_custom/villani_plus_abud_sce.rds')
## table(se$treatment)

## The following list of commands will generate the plots created in iSEE
## Copy them into a script or an R session containing your SingleCellExperiment.
## All commands below refer to your SingleCellExperiment object as `se`.

colormap <- ExperimentColorMap()
colormap <- synchronizeAssays(colormap, se)
all_contents <- list()

################################################################################
## Complex heatmap 1
################################################################################


.chosen.rows <- c("ENSG00000204531.21__POU5F1", "ENSG00000111704.11__NANOG", "ENSG00000156574.10__NODAL", 
    "ENSG00000241186.10__TDGF1", "ENSG00000174720.18__LARP7", "ENSG00000156925.12__ZIC3", 
    "ENSG00000133980.5__VRTN", "ENSG00000136110.13__CNMD", "ENSG00000147869.5__CER1", 
    "ENSG00000184344.4__GDF3", "ENSG00000121570.13__DPPA4", "ENSG00000240563.2__L1TD1", 
    "ENSG00000069482.7__GAL", "ENSG00000147601.15__TERF1", "ENSG00000126010.6__GRPR", 
    "ENSG00000143768.13__LEFTY2", "ENSG00000086475.15__SEPHS1", "ENSG00000141682.12__PMAIP1", 
    "ENSG00000095002.15__MSH2", "ENSG00000171634.19__BPTF", "ENSG00000132330.18__SCLY", 
    "ENSG00000144401.14__METTL21A", "ENSG00000111752.11__PHC1", "ENSG00000174059.17__CD34", 
    "ENSG00000081237.21__PTPRC", "ENSG00000197471.12__SPN", "ENSG00000169313.10__P2RY12", 
    "ENSG00000204472.13__AIF1", "ENSG00000066336.12__SPI1",  
    "ENSG00000095970.17__TREM2", "ENSG00000169896.18__ITGAM", "ENSG00000130203.10__APOE", 
    "ENSG00000153208.18__MERTK", "ENSG00000082781.12__ITGB5", "ENSG00000168329.14__CX3CR1", 
    "ENSG00000106799.14__TGFBR1", "ENSG00000184500.16__PROS1", "ENSG00000171659.16__GPR34", 
    "ENSG00000134508.13__CABLES1", "ENSG00000123095.6__BHLHE41", "ENSG00000116774.12__OLFML3", 
    "ENSG00000137491.15__SLCO2B1", "ENSG00000092068.21__SLC7A8", "ENSG00000112033.14__PPARD", 
    "ENSG00000100122.7__CRYBB1", "ENSG00000103066.13__PLA2G15", "ENSG00000117984.15__CTSD", 
    "ENSG00000135838.14__NPL", "ENSG00000134575.13__ACP2", "ENSG00000135047.16__CTSL", 
    "ENSG00000135124.16__P2RX4", "ENSG00000100997.20__ABHD12", "ENSG00000100985.7__MMP9", 
    "ENSG00000108691.10__CCL2", "ENSG00000133048.13__CHI3L1", "ENSG00000106991.14__ENG", 
    "ENSG00000159189.13__C1QC", "ENSG00000173369.18__C1QB", "ENSG00000173372.18__C1QA", 
    "ENSG00000151117.9__TMEM86A", "ENSG00000163106.11__HPGDS", "ENSG00000282608.2__ADORA3", 
    "ENSG00000143344.16__RGL1", "ENSG00000153071.15__DAB2", "ENSG00000275302.2__CCL4", 
    "ENSG00000010327.10__STAB1", "ENSG00000124225.16__PMEPA1", "ENSG00000115008.6__IL1A", 
    "ENSG00000107719.10__PALD1", "ENSG00000167601.12__AXL", "ENSG00000244405.8__ETV5", 
    "ENSG00000139572.4__GPR84", "ENSG00000171631.16__P2RY6", "ENSG00000112486.18__CCR6", 
    "ENSG00000205336.14__ADGRG1", "ENSG00000213316.10__LTC4S", "ENSG00000148908.15__RGS10", 
    "ENSG00000159216.19__RUNX1", "ENSG00000182578.14__CSF1R", "ENSG00000183087.15__GAS6", 
    "ENSG00000105383.15__CD33", "ENSG00000120738.8__EGR1", "ENSG00000064687.13__ABCA7", 
    "ENSG00000085514.16__PILRA", "ENSG00000174837.15__ADGRE1", "ENSG00000163464.8__CXCR1", 
    "ENSG00000140678.17__ITGAX", "ENSG00000175445.17__LPL");

```


```{r}
## now, sort the columns as requested
##   1.  hiPSC (Abud)
##   2.  hiHPC (Abud)
##   3.  Blood DC (Abud)
##   4.  CD14 M (Abud)
##   5.  CD16 M (Abud)
##   6.  Etal MG (Abud)
##   7.  Adult MG (Abud)
##   8.  hiPSC-MG (Abud)
## 9.  hiPSC-MG (this work)
## plus mutants


sorted_cols <- c( "iPSC_(Abud)",  "iHPC_(Abud)", "Blood_DC_(Abud)" ,    
                 "CD14_M_(Abud)",   "CD16_M_(Abud)",
                 "Fetal_MG_(Abud)",  "Adult_MG_(Abud)",
                 "iPSC_MG_(Abud)", "iPSC_MG",
                 "iPSC_MG_Slc37a2_mut1", "iPSC_MG_Slc37a2_mut2",    "iPSC_MG_TREM2_mut") 



```



```{r}
.chosen.columns <- colnames(se);
plot.data <- assay(se, "logcpms")[.chosen.rows, .chosen.columns, drop=FALSE]
plot.data <- as.matrix(plot.data);

plot.data <- plot.data - rowMeans(plot.data)
plot.data <- plot.data / apply(plot.data, 1, sd)

.assay_colors <- c("blue", "white", "red")
.assay_colors <- circlize::colorRamp2(breaks = c(-3.73423928357544, 0, 3.44073249010621), colors = .assay_colors)

# Keep all samples to compute the full range of continuous annotations
.column_data <- colData(se)[, c("origin", "treatment"), drop=FALSE]

.column_col <- list()

.column_data[["origin"]] <- factor(.column_data[["origin"]]);
.color_values <- .column_data[["origin"]]
.color_values <- setdiff(unique(.color_values), NA)
.col_colors <- colDataColorMap(colormap, "origin", discrete=TRUE)(length(.color_values))
names(.col_colors) <- unique(.color_values)
.column_col[["origin"]] <- .col_colors

.column_data[["treatment"]] <- factor(.column_data[["treatment"]]);
.color_values <- .column_data[["treatment"]]
.color_values <- setdiff(unique(.color_values), NA)
.col_colors <- colDataColorMap(colormap, "treatment", discrete=TRUE)(length(.color_values))
names(.col_colors) <- unique(.color_values)
.column_col[["treatment"]] <- .col_colors

.column_data <- .column_data[colnames(plot.data), , drop=FALSE]
.column_data <- as.data.frame(.column_data, optional=TRUE)
.column_annot_order <- order(.column_data[["origin"]], .column_data[["treatment"]])
.column_data <- .column_data[.column_annot_order, , drop=FALSE]
plot.data <- plot.data[, .column_annot_order, drop=FALSE]
.column_annot <- ComplexHeatmap::columnAnnotation(df=.column_data, col=.column_col, annotation_legend_param=list(direction="vertical"))
```

Shortened and sorted gene names:

```{r}
rownames(plot.data) <- sapply(strsplit(rownames(plot.data), '__'), function(x) return(x[2]))
plot.data <- plot.data[c(pluri, hpc, micro),]
row.split <- data.frame(gene = c(pluri, hpc, micro),
                        category = NA)
row.split[row.split$gene %in% pluri, 'category'] <- 'Pluripotency'
row.split[row.split$gene %in% hpc, 'category'] <- 'HPC'
row.split[row.split$gene %in% micro, 'category'] <- 'Microglia'


rowha <- rowAnnotation(signature = row.split$category,
                       show_annotation_name = TRUE)

hm <- ComplexHeatmap::Heatmap(matrix=plot.data, col=.assay_colors,
                              top_annotation=.column_annot, cluster_rows=TRUE,
                              split = as.factor(row.split$category),
                              left_annotation = rowha,
                              clustering_distance_rows="spearman", clustering_method_rows="ward.D2",
                              cluster_columns=FALSE, name="logcpms\n(centered, scaled)",
                              show_row_names=TRUE, show_column_names=TRUE,
                              heatmap_legend_param=list(direction="vertical"))

```


```{r heatmap_mutants, fig.width = 16, fig.height = 16, results = 'as-is'}
ComplexHeatmap::draw(hm, heatmap_legend_side="right", annotation_legend_side="right")

```




# Trace

```{r}
sessionInfo()
```

