---
title: "Ambra Villani's microglia RNA-seq, custom plots"
author: "Izaskun Mallona, Mark D. Robinson's lab, UZH"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: kable
    theme: lumen
params:
  seed: 665
---



```{r warning = FALSE, message = FALSE}

library(ggplot2)
library(SingleCellExperiment)
library(Cairo)
library(scater)
library(iSEE)
library(ComplexHeatmap)

## options(bitmapType="cairo")
```

```{r}
## render on error
knitr::knit_hooks$set(error = function(x, options) {
    knitr::knit_exit()  
})
```

```{r}
knitr::opts_chunk$set(fig.width = 5,
                      fig.height = 5,
                      cache = FALSE,
                      include = TRUE,
                      fig.path = "other_plots/",
                      dev = c('svg', "png"),
                      cache.lazy = FALSE,
                      warning = TRUE,
                      message = TRUE)

## plan("multiprocess", workers = NTHREADS)
## options(future.globals.maxSize= 1.4e9)

## options(bitmapType='cairo')
```

# Note

To be knitted with `/usr/bin/R`


# Aim 1 {.tabset .tabset-pills}

Updating of the original request (last chunk) as here:

<details>
Nothing major and no re-analysis needed: during review we decided to remove TMEM119 to avoid overinterpretation, so we would simply need to update the RNA-seq plots accordingly.

Concretely, this concerns:

Supplementary Figure 1C and 1D: removal of TMEM119 from the gene lists / plots.

</details>

Custom re-plotting, by Ambra:

<details>

```
Plot you did last time (WT data, by replicate, also attached here) with the following edits:
Remove gene Slc37a2
Show only the time points 34 and 38 (remove 28)
```

</details>


## Symbol depicts timepoint


```{r}
se <- readRDS('../03_custom/ambra_villani_only_sce.rds')


se <- se[,se$mutation == 'WT']
se <- se[,se$timepoint %in% c(34, 38)]

targets <- read.table(text='ENSG00000169313.10__P2RY12
ENSG00000066336.12__SPI1
ENSG00000095970.17__TREM2
ENSG00000169896.18__ITGAM
ENSG00000130203.10__APOE
ENSG00000153208.18__MERTK
ENSG00000168329.14__CX3CR1
ENSG00000159189.13__C1QC
ENSG00000173369.18__C1QB
ENSG00000173372.18__C1QA
ENSG00000171631.16__P2RY6
ENSG00000183087.15__GAS6
ENSG00000175445.17__LPL
ENSG00000204472.13__AIF1
ENSG00000141458.13__NPC1
ENSG00000119655.11__NPC2
ENSG00000181631.7__P2RY13')$V1

targets_short <- sapply(strsplit(targets, '__'), function(x) return(x[2]))
cols <- c('WT' = '#33CC66')
```

```{r bytimepoint_wt, fig.width = 6.1, fig.height = 4}

foo <- scater::plotExpression(se, targets, 
               shape_by = "timepoint",
               colour_by = 'mutation',
               exprs_values = 'logcpm')

foo +
    scale_x_discrete(labels = targets_short) +
    scale_colour_manual(name = 'background', values = cols) +
    labs(shape = 'timepoint') 
```

## Symbol depicts replicate


```{r byreplicate_wt, fig.width = 7, fig.height = 4}

colData(se)$replicate <- colData(se)$treatment

foo <- scater::plotExpression(se, targets, 
               shape_by = "replicate",
               colour_by = 'mutation',
               exprs_values = 'logcpm')

foo +
    scale_x_discrete(labels = targets_short) +
    scale_colour_manual(name = 'background', values = cols)

```


# Aim 2 {.tabset .tabset-pills}

Following Ambra's:

<details>

```
A new plot Including:
 - Backgrounds: WT, Slc37a2, TREM2
- Show for all only time point 28 (not used for the first plot)
- Use the same list of genes as the one attached here (including Slc37a2), following this order:
  - Position 1: Slc37a2
  - Position 2: TREM2
  - The position of the next genes doesn’t matter!
```
</details>


## Symbols depict timepoint


```{r}
se <- readRDS('../03_custom/ambra_villani_only_sce.rds')

se <- se[,se$timepoint == 28]

targets <- read.table(text='
ENSG00000134955.12__SLC37A2
ENSG00000095970.17__TREM2
ENSG00000169313.10__P2RY12
ENSG00000066336.12__SPI1
ENSG00000169896.18__ITGAM
ENSG00000130203.10__APOE
ENSG00000153208.18__MERTK
ENSG00000168329.14__CX3CR1
ENSG00000159189.13__C1QC
ENSG00000173369.18__C1QB
ENSG00000173372.18__C1QA
ENSG00000171631.16__P2RY6
ENSG00000183087.15__GAS6
ENSG00000175445.17__LPL
ENSG00000204472.13__AIF1
ENSG00000141458.13__NPC1
ENSG00000119655.11__NPC2
ENSG00000181631.7__P2RY13')$V1

targets_short <- sapply(strsplit(targets, '__'), function(x) return(x[2]))
cols <- c('WT' = '#33CC66', 'Slc37a2' = '#E69F00', 'TREM2' = '#0072B2')
```


```{r bytimepoint_mutants, fig.width = 6.1, fig.height = 4}

foo <- scater::plotExpression(se, targets, 
               shape_by = "timepoint",
               colour_by = 'mutation',
               exprs_values = 'logcpm')

foo +
    scale_x_discrete(labels = targets_short) +
    scale_colour_manual(name = 'background', values = cols) +
    labs(shape = 'timepoint') 
```

## Symbol depicts replicate


```{r byreplicate_mutants, fig.width = 7, fig.height = 4}

colData(se)$replicate <- colData(se)$treatment

foo <- scater::plotExpression(se, targets, 
               shape_by = "replicate",
               colour_by = 'mutation',
               exprs_values = 'logcpm')

foo +
    scale_x_discrete(labels = targets_short) +
    scale_colour_manual(name = 'background', values = cols)

```


# Trace

```{r}
sessionInfo()
```

